\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[all]{xy}
\usepackage{color}
\usepackage{geometry}
\usepackage{hyperref}

\makeatletter
\renewcommand*\env@matrix[1][*\c@MaxMatrixCols c]{%
  \hskip -\arraycolsep
  \let\@ifnextchar\new@ifnextchar
  \array{#1}}
\makeatother
\makeatletter
\DeclareRobustCommand{\pns}{\mathrel{\text{$\m@th\proper@ideal$}}}
\newcommand{\proper@ideal}{%
  \ooalign{$\lneq$\cr\raise.22ex\hbox{$\lhd$}\cr}%
}
\makeatother

\SelectTips{eu}{}
\setlength{\fboxsep}{0pt}
\setlength\parskip{0.3em}
\setlength{\parindent}{0 pt}

\newcommand{\la}{\left\langle}
\newcommand{\ra}{\right\rangle}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Dic}{\text{Dic}}
\newcommand{\Aut}{\text{Aut}}
\newcommand{\Syl}{\text{Syl}}
\newcommand{\orb}{\text{orb}}
\newcommand{\stab}{\text{stab}}
\newcommand{\fix}{\text{fix}}
\newcommand{\Sym}{\text{Sym}}
\newcommand{\Alt}{\text{Alt}}
\newcommand{\supp}{\text{supp}}
\newcommand{\adj}{\text{adj}\ }
\newcommand{\lcm}{\text{lcm}\ }
\newcommand{\id}{\text{id}}
\newcommand{\im}{\text{im\,}}
\newcommand{\spanset}{\text{span}}
\newcommand{\rank}{\text{rank }}
\newcommand{\Mod}{\text{ mod }}

\theoremstyle{definition}

\newtheorem{defn}{Definition}[subsection]
\newtheorem{prop}[defn]{Proposition}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{coro}[defn]{Corollary}
\newtheorem{example}[defn]{Example}
\newtheorem{exe}[defn]{Exercise}
\newtheorem{tool}[defn]{Tool}
\newtheorem*{remark}{Remark}
\newtheorem*{notation}{Notation}

\title{MA3K4 Introduction to group theory :: Lecture notes}
\author{Lecturer: Gareth Tracey}
\date{\today}

\begin{document}

\maketitle
\thispagestyle{empty}

\tableofcontents
\thispagestyle{empty}
\newpage
\setcounter{page}{1}

\begin{flushright}
\textit{Week 1, lecture 1 starts here}
\end{flushright}

\section{Introduction}
\begin{defn}
A \textit{group} is a pair $(G,\circ)$ where $G$ is a set and $\circ:G\times G\rightarrow G$ is a binary operation satisfying
\begin{enumerate}
\item Associativity: $(g\circ h)\circ k = g\circ (h\circ k) \ \forall g,h,k \in G$,
\item Identity: $\exists$ an element in $G$, denoted $1_G$, such that $1_G\circ g = g\circ 1_G = g \ \forall g\in G$,
\item Inverses: $\forall g\in G$, $\exists$ an element in $G$, denoted $g^{-1}$, such that $g\circ g^{-1} = g^{-1}\circ g = 1_G$.
\end{enumerate}
\end{defn}

\begin{remark}
Implicit in parts 1 and 2 of above definition are
\begin{enumerate}
\item An identity element in an associative binary operation is unique, justifying the notation and the `the' before `identity'
\item Similarly, inverses are unique in an associative binary operation, so we say \textit{the} inverse of $g$
\end{enumerate}
The number of elements in a group $(G,\circ)$ is called the order of $G$, denoted $|G|$.
\end{remark}

\begin{example}
Let $G=\Z$. Then
\begin{enumerate}
\item If we define $\circ:G\times G\rightarrow G$ by $g\circ h=g+h$ for $g,h\in \Z$ then we know $(G,\circ)$ is a group and $1_G=0,\ g^{-1}=-g \ \forall g\in G$.
\item For the same set, if we define $g\circ h=g\times h$ then $(G,\circ)$ is not a group for lack of inverses for $g\in \Z \backslash \{\pm 1\}$.
\end{enumerate}
\end{example}

\begin{remark}
\begin{enumerate}
\item You may have been given a fourth axiom, closure, in previously seen definitions of a group. The reason we omit that here is because it's implied by definition of binary operation.
\item If $(G,\circ)$ is a group, $\circ$ is often called the \textit{group operation}.
\item Given clear context, we will streamline our notation and simply write $G$ in place of $(G,\circ)$ and $gh$ in place of $g\circ h$.
\end{enumerate}
\end{remark}

\begin{defn}
Let $G$ be a group.
\begin{enumerate}
\item If $g,h\in G : gh=hg$ then $g$ and $h$ \textit{commute}.
\item If $g$ and $h$ commute $\forall g,h \in G$ then $G$ is \textit{abelian}.
\end{enumerate}
\end{defn}

\begin{example}
$(\Z,+)$ is abelian.
\end{example}

\begin{exe}[Commuting elements in groups]
Let $G$ be a group.
\begin{enumerate}
\item Suppose $g^2=1_G \ \forall g\in G$. Show that $G$ is abelian.
\begin{proof}
Note that this implies $\forall g,h\in G,\ (gh)^{-1}=gh$, but $(gh)^{-1}=h^{-1}g^{-1}=hg$, so $gh=hg$.
\end{proof}
\item Suppose $g^3=1_G \ \forall g\in G$. Show that $hgh^{-1}$ and $g$ commute $\forall g,h\in G$.
\begin{proof}
One has $g^2h=g^{-1}h^{-2}=(h^2g)^{-1}=h^2gh^2g\Rightarrow gh^2g=hg^2h\Rightarrow  hgh^2g=h^2g^2h$. Now consider $(gh)^{-1}$, which equals $h^2g^2$ but also $ghgh$. Hence $ghgh^{-1}=ghgh^2=h^2g^2h=hgh^2g=hgh^{-1}g$, as desired.
\end{proof}
\end{enumerate}
\end{exe}

Next, we are going to look at two infinite families of examples of groups: 1. Symmetric groups and 2. Linear groups.

\subsection{Symmetric group}
\begin{defn}
Let $X$ be a set, and define
\[
\Sym(X) = \{f:f:X\rightarrow X \text{ is a bijection}\}
\]
Define $\circ:\Sym(X)\times \Sym(X)\rightarrow \Sym(X)$ to be the usual composition of functions. Then $(\Sym(X),\circ)$ is a group, called the \textit{symmetric group} on $X$. An element of $\Sym(X)$ is called a \textit{permutation}. 
\end{defn}

\begin{remark}[Sanity check]
\begin{enumerate}
\item Associativity is clear by inheritance
\item $1_G=\id_X : x\mapsto x$
\item For $f\in \Sym(X)$, $x\in X$, choose a unique $y_x\in X$ such that $f(y_x)=x$. Define $g:X\rightarrow X$ by $g(x)=y_x$, then $g$ is a inverse for $f$.
\end{enumerate}
\end{remark}

We introduce cycle notation as a more compact way of writing permutations down.

\begin{flushright}
\textit{Week 1, lecture 2 starts here}
\end{flushright}

\begin{defn}[Cycle notation]
Let $X$ be a set.
\begin{enumerate}
\item Let $a_1,\ldots,a_n\in X$ be distinct. The permutation $f=(a_1,\ldots,a_n) \in\Sym(X)$ is defined to be $f(a_i)=a_{i+1}$ for $1\leq i\leq n-1$, $f(a_{n})=a_1$, and $f(b)=b$ for $b\not\in \{a_1,\ldots,a_n\}$. We call $f$ a \textit{cycle of length} $n$ (or an $n$-cycle).
\item Two cycles $(a_1,\ldots,a_r),\ (b_1,\ldots,b_s)$ are \textit{disjoint} if $\{a_1,\ldots,a_r\} \cap \{b_1,\ldots,b_s\}=\varnothing$.
\item The \textit{empty cycle}, written $()$, is the identity map which is also $1_{\Sym(X)}$.
\end{enumerate}
\end{defn}
\begin{remark}[Important points about cycles]
\begin{enumerate}
\item Perhaps a tautology, but the empty cycle is thought of as a cycle (of length 0).
\item Recall that the group operation is composition of functions. So $fg:X\rightarrow X$ means do $g$ first and then $f$. e.g. $X=\{1,2,3,4,5\}$, so $(3,4,1,2)(4,5)=(1,2,3,4,5)$.
\item Cycle notation is not unique in the following sense: two distinct $m$-tuples of elements in a set $X$ can represent the same cycle, e.g. $(1,2,3,4,5)=(3,4,5,1,2)$.
\end{enumerate}
\end{remark}

\begin{thm}
Let $X$ be a finite set. Then
\begin{enumerate}
\item $|\Sym(X)|=|X|!$,
\item Every element $F\in \Sym(X)$ can be written as product of disjoint cycles. Moreover, the decomposition is unique in the sense that if $F=f_1\cdots f_r=g_1\cdots g_s$ where $f_i,g_i$ are disjoint cycles of length $>1$, then $r=s$ and $\{f_1,\ldots,f_r\}=\{g_1,\ldots,g_r\}$.
\end{enumerate}
\end{thm}
\begin{proof}[Proof (nonexaminable)]
\begin{enumerate}
\item Write $X=\{x_1,\ldots,x_r\}$ where $n=|X|$ and define
\[
X(n):=\{(a_1,\ldots,a_n):a_i\in X, a_i\neq a_j \text{ for } i\neq j\}.
\]
Define a bijection $\theta:\Sym(X)\rightarrow X(n)$ by $\theta(f)=(f(x_1),\ldots,f(x_n))$. for $f\in \Sym(X)$, observe
\begin{enumerate}
\item $\theta$ is well-defined, since $f$ is a bijection, so $f(x_i)\neq f(x_j)$ for $i\neq j$.
\item In the same way, $\theta$ is injective. Indeed, if $\theta(f)=\theta(g)$ then $f(x_i)=g(x_i) \ \forall i$ by definition of $\theta$, so $f=g$.
\item If $(a_1,\ldots,a_n)\in X(n)$, then define $f:X\rightarrow X$ by $f(x_i)=a_i$ for $1\leq i\leq n$. Clearly, $f\in \Sym(X)$ and $\theta(f) = (a_1,\ldots,a_n)$, so $\theta$ is surjective.
\end{enumerate}
It follows that $|\Sym(X)|=|X(n)|=n!$.
\item Let $f\in\Sym(X)$. If $f=\id_X$ then $f=()$ so it's a cycle. Now suppose $f$ is not $\id_X$. Let $Y=\{x\in X:f(x)\neq x\}$. Note that since $|\Sym(X)|$ is finite by 1., $\exists n\in \N$ such that $f^n=\id_X$.

In particular, if we fix $a_1\in Y$, then we may define $m_1:=\min\{m\in \N:f^m(a_1)=a_1\}$ since the set is nonempty. Now, for $2\leq i\leq m_1$, define $a_i:=f(a_{i-1})$. If $Y=\{a_1,\ldots,a_{m_1}\}$, then by definition of cycle, one has $f=(a_1,\ldots,a_m)$.

Now suppose $Y\backslash \{a_1,\ldots,a_{m_1}\}\neq \varnothing$. Choose $a_{m_1+1} \in Y\backslash \{a_1,\ldots,a_{m_1}\}$, and define $m_2:=\min\{m\in \N:f^m(a_{m_1+1})=a_{m_1+1}\}$. For $m_1+2\leq i\leq m_2$, again define $a_i:=f(a_{i-1})$, then if $Y=\{a_1,\ldots,a_{m_1},a_{m_1+1},\ldots,a_{m_2}\}$, one has $f=(a_1,\ldots,a_m)(a_{m+1},\ldots,a_{m_2})$. If not, we continue inductively. Since $X$ is finite, this must terminate, and when it does $f$ will be a product of disjoint cycles. The uniqueness follows from the algorithm immediately.
\end{enumerate}
\end{proof}

\subsection{Linear group}
\begin{defn}
$F$ is a field and $n\in \N$. We define
\[
GL_n(F) := \{A:A\text{ an invertible }n\times n \text{ matrix over } F\},
\]
a group with matrix multiplication as operation. This is called \textit{general linear group} of dimension $n$ over $F$.
\end{defn}
\begin{flushright}
\textit{Week 1, lecture 3 starts here}
\end{flushright}

\begin{remark}[Useful things from Algebra I, II for studying general linear groups]
\begin{enumerate}
\item Each field $F$ has an additive and multiplicative identity $0_F$ and $1_F$. Given clear context, they will be denoted simply 0 and 1 respectively.
\item An $n\times n$ matrix $A$ over $F$ is invertible iff $\det A\neq 0$ iff rows (or columns) of $A$ are linearly independent.
\item If $F$ is a finite field, then $|F|=p^f$ for some prime $p$ and $f\in\N$. Moreover, for each prime $p$ and each $f\in\N$, $\exists!$ a field (up to isomorphism) $F:|F|=p^f$. $p$ is called the \textit{characteristic} of $F$, and satisfies that $p\alpha=0 \ \forall \alpha\in F$.
\item If $F$ is a field then $F^{\times}:=F\backslash \{0\}$ is a group with multiplication as group operation inherited from $F$.
\end{enumerate}
\end{remark}
\begin{exe}
\begin{enumerate}
\item Let $X$ be a set. Show that $\Sym(X)$ is abelian iff $|X|\leq 2$.
\item Let $F$ be a field. Show that $GL_n(F)$ is abelian iff $n=1$.
\end{enumerate}
\end{exe}
\begin{thm}
\label{thm:orderofGLnF}
Let $F$ be a finite field with $|F|=q$. Then $\displaystyle |GL_n(F)|=q^{\binom{n}{2}} \prod_{i=1}^n (q^i-1)$.
\end{thm}
\begin{proof}[Proof (nonexaminable)]
See sheet 1. 
\end{proof}

\subsection{Order of elements}
\begin{defn}
The \textit{order} of $g\in G$, denoted $|g|$, is defined $|g|:=\min \{n\in\N:g^n= 1_G\}$. If the set is $\varnothing$ then $|g|:=\infty$.
\end{defn}
\begin{example}
\begin{enumerate}
\item Let $X$ be a set and let $f=(a_1,\ldots,a_m)\in\Sym(X)$. Then $|f|=m$.
\item Let $F$ be a finite field of order $p^f$ where $p$ prime, $G=GL_2(F)$, and $\alpha,\beta\in F^{\times}$. Observe that
\[
\begin{pmatrix}
1 & \alpha \\ 0 & 1
\end{pmatrix}
\begin{pmatrix}
1 & \beta \\ 0 & 1
\end{pmatrix}=
\begin{pmatrix}
1 & \alpha+\beta \\ 0 & 1.
\end{pmatrix}
\]
So if $g=\begin{pmatrix}
1 & \alpha \\ 0 & 1
\end{pmatrix}$ then $g^n=\begin{pmatrix}
1 & n\alpha \\ 0 & 1
\end{pmatrix}$, so $|g|\mid p$ (we'll see later about this implication), so $|g|=p$.

Also,
\[
g^n=\begin{pmatrix}
\alpha & 0 \\ 0 & \beta
\end{pmatrix}^n=
\begin{pmatrix}
\alpha^n & 0 \\ 0 & \beta^n .
\end{pmatrix}
\]
So $|g|=\lcm (m,k)$ where $m=|\alpha|$ and $k=|\beta|$ as elements of $F^\times$.
\end{enumerate}
\end{example}
\begin{remark}
\begin{enumerate}
\item For $g\in G$, $(g^n)^{-1}=\left(g^{-1}\right)^n$, so we write $g^{-n}:=\left(g^{-1}\right)^n$. In particular, $\left|g^{-1}\right|=|g|$.
\item If $g\in G,\ n=|g|$ and $n\mid l$, then $g^l=1$.
\end{enumerate}
\end{remark}
\begin{lemma}
\label{lemma:orderlem}
Let $a,b\in G$ of finite order. Then
\begin{enumerate}
\item If $l\in \N$, then $a^l=1$ iff $|a|\mid l$.
\item Let $m\in \N$, then $\displaystyle |a^m|=\frac{|a|}{\gcd (|a|,m)}$.
\item If $a,b$ commute then $|ab|\mid \lcm (|a|,|b|)$.
\item If $a,b$ commute and $a^i=b^j \ \forall i,j\in \N$ only when they are both 1 (i.e. $\la a\ra \cap \la b\ra =\{1\}$) then $|ab|= \lcm (|a|,|b|)$.
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{enumerate}
\item $\Leftarrow$ is mentioned. $\Rightarrow$: suppose $a^l=1$. By Euclidean division, we can write $l=q|a|+r$ for some $r\in[0,|a|)$. Then $1=a^l=a^{q|a|+r}=a^r$, which contradicts minimality of $|a|$.
\item Suppose first that $m\mid |a|$. Then one can write $|a|=ms$, so $a^{ml}=1\Leftrightarrow |a| \mid ml$ by 1 $\displaystyle \Leftrightarrow \frac{|a|}{m}\mid l$. Hence the least positive integer $l:a^{ml}=1$ is $\displaystyle \frac{|a|}{m}$.

Now let $k=\gcd (|a|,m)$. We write $m=ks$, then $a^{m\frac{|a|}{k}}=a^{|a|s}=1$, and by 1 one has $|a^m| \mid \frac{|a|}{k}$. To complete the proof it suffices to show that $\frac{|a|}{k}\leq |a^m|$.
\begin{flushright}
\textit{Week 2, lecture 1 starts here}
\end{flushright}
By Bézout's lemma, $\exists s,t\in\Z:k=s|a|+tm$, so $a^k=a^{s|a|+tm}=(a^{|a|})^s a^{tm}=a^{tm}$. Then $a^{tm|a^m|}=((a^m)^{|a^m|}))^t=1^t=1$. This implies $|a^{tm}|\mid a^m$ by 1. So $\frac{|a|}{k}=|a^k|=|a^tm|\mid |a^m|$.
\item Let $l:=\lcm(|a|,|b|)$. Then $(ab)^l=a^lb^l=1\times 1=1$, so by 1. $|ab|\mid l$.
\item Let $k:=|ab|$. Then $k\mid l$, but also, $1=(ab)^k=a^kb^k$ so $a^k=\left(b^{-1}\right)^k$ and by assumption both sides are 1. So $|a|,|b|\mid k$, so $l\mid k$, hence $k=l$.
\end{enumerate}
\end{proof}
\begin{exe}
\label{exe:conjaugatesordersequal}
\begin{enumerate}
\item Let $h,g\in G$. Show that $\left|hgh^{-1}\right|=|g|$.
\item Let $l,m,n>2\in\N$. Show that $\exists G$ with $a,b\in G:|a|=l,\ |b|=m,\ |ab|=n$. Also:
\begin{enumerate}
\item Show that $G$ can be finite.
\item Show that one can replace $l,m,n>2$ by $l,m,n>1$.
\end{enumerate}
Key hint: A $2\times 2$ matrix over $\mathbb C$ with distinct eigenvalues is diagonalisable. Now exploit result of 1st exercise.
\end{enumerate}
\end{exe}
\subsection{Subgroup and coset}
\begin{defn}
A nonempty $H\subseteq G$ is a \textit{subgroup} of $G$, denoted $H\leq G$, if
\begin{enumerate}
\item $1_G\in H$
\item $h\in H\Rightarrow h^{-1}\in H$
\item $h_1,h_2\in H \Rightarrow h_1h_2\in H$
\end{enumerate}
\end{defn}
\begin{defn}
For a group $G$ and $g\in G$, define $\la g\ra:=\{g^n:n\in\Z\}$ which is called the \textit{cyclic subgroup of} $G$ \textit{generated by} $g$. If $G=\la g\ra$ then $G$ is \textit{cyclic} and $g$ is a \textit{generator} for $G$.
\end{defn}
\begin{lemma}
\label{lemma:sgptest}
$H\subseteq G$ where $H$ nonempty. $H\leq G \Leftrightarrow h_1,h_2\in H\Rightarrow h_1h_2^{-1}\in H$
\end{lemma}
\begin{proof}
\begin{itemize}
\item[$\Rightarrow$] $h_1,h_2\in H\Rightarrow h_2^{-1}\in H\Rightarrow h_1h_2^{-1}\in H$.
\item[$\Leftarrow$] \begin{enumerate}
\item $H\neq \varnothing\Rightarrow h\in H\Rightarrow hh^{-1}\in H\Rightarrow 1_G\in H$
\item $h\in H\Rightarrow 1_Gh^{-1}=h^{-1}\in H$
\item $h_1,h_2\in H\Rightarrow h_2^{-1}\in H\Rightarrow h_1(h_2^{-1})^{-1}h_1h_2\in H$
\end{enumerate}
\end{itemize}
\end{proof}
\begin{example}
Let $G=GL_2(F)$ and
\[
H=\left\{\begin{pmatrix}
  \alpha & 0 \\ 0 & \beta
\end{pmatrix}:\alpha,\beta\in F^{\times}\right\} \subseteq G. \qquad\text{sometimes called diagonal subgroup}
\]
We want to show this is indeed a subgroup. Let $h_i=\begin{pmatrix}
  \alpha_i & 0 \\ 0 & \beta_i
\end{pmatrix}\in H$ where $i=1,2$. Then
\[
h_1h_2=\begin{pmatrix}
  \alpha_1 & 0 \\ 0 & \beta_2
\end{pmatrix}\begin{pmatrix}
  \alpha_2^{-1} & 0 \\ 0 & \beta_2^{-1}
\end{pmatrix}=\begin{pmatrix}
  \alpha_1 \alpha_2^{-1} & 0 \\ 0 & \beta_1\beta_2^{-1}
\end{pmatrix}\in H .
\]
\end{example}
\begin{defn}
Let $A\subseteq G$ be nonempty. The \textit{subgroup of} $G$ \textit{generated by} $A$, denoted $\la A\ra$, is
\[
\{a_1^{\varepsilon_1}\cdots a_m^{\varepsilon_m}:m\in\N,\ a_i\in A,\ \varepsilon_i\in\{\pm 1\}\}.
\]
\end{defn}
\begin{notation}
If $A=\{g_1,\ldots,g_t\}$ then we often write $\la A\ra$ as $\la g_1,\ldots,g_t\ra$.
\end{notation}

\begin{flushright}
\textit{Week 2, lecture 2 starts here}
\end{flushright}

\begin{exe}
Let $G$ be a group and $A\subseteq$ nonempty.
\begin{enumerate}
\item Use Lemma \ref{lemma:sgptest} to show that $\la A\ra$ is indeed a subgroup of $G$.
\item Write $A=\{g_1,\ldots,g_s\}$ and suppose $g_ig_j=g_jg_i \ \forall i,j=1,\ldots,s$. Show that $|\la A\ra|\leq \prod_{i=1}^s |g_i|$.
\item Suppose $g^p=1 \ \forall g\in G$ and $G=\la x,y\ra$ for some $x,y\in G$.
\begin{enumerate}
\item Show that if $p=2,\ |G|\leq 4$.
\item Show that if $p=3,\ |G|\leq 3^4$.
\item Fields-medal-worth: If $p=5$, is $G$ finite?
\end{enumerate}
\end{enumerate}
\end{exe}

\begin{defn}
The \textit{left coset} of $H\leq G$ with respect to $g\in G$ is the set $gH:=\{gh:h\in H\}.$ The \textit{right coset} is defined similarly.
\end{defn}
$gH$ is not a subgroup unless $g\in H$ since in general the identity is not there.
\begin{lemma}
Let $H\leq G$ and $g,k\in G$. The following are equivalent:
\begin{enumerate}
\item $k\in gH$
\item $kH=gH$
\item $g^{-1}k\in H$
\end{enumerate}
\end{lemma}
\begin{proof}
First note that if $h\in H$ then $hH=H$ by virtue of the fact $H\leq G$.

Now $k\in gH\Rightarrow k=gh$ for some $h\in H \Rightarrow kH=ghH=gH$, so 1 implies 2. The other two implications are almost identical.
\end{proof}
\begin{lemma}
Let $H\leq G$. For $g_1,g_2\in G$, say that $g_1\sim_H g_2\Leftrightarrow g_1H=g_2H$. Then $\sim_H$ is an equivalence relation.
\end{lemma}
\begin{proof}
The three conditions reflexivity, symmetry and transitivity follow immediately from definition.
\end{proof}
\begin{coro}
\label{coro:cosetspartitiongp}
Let $H\leq G$.
\begin{enumerate}
\item If $g_1,g_2\in G$, then either $g_1H=g_2H$ or $g_1H\cap g_2 H=\varnothing$.
\item The set $\{gH:g\in G\}$ of left cosets is a partition of $G$, i.e. if $g_iH$ for $i\in I$ are distinct left cosets of $H$ in $G$ then
\[
G=\bigsqcup_{i\in I} g_i H.
\]
\end{enumerate}
\end{coro}
\begin{proof}
$\{gH:g\in G\}$ is precisely the set of equivalence classes under $\sim_H$, so the results follow immediately.
\end{proof}
\begin{thm}[Lagrange's]
Let $G$ be a finite group and $H\leq G$. Then $|H|\mid |G|$.
\end{thm}
\begin{proof}
Let $g_1H,\ldots,g_tH$ be distinct left cosets of $H$ in $G$. By Corollary \ref{coro:cosetspartitiongp},
\[
|G|=\left|\bigsqcup_{i=1}^t g_i H\right|=\sum_{i=1}^t |g_iH|,
\]
and one also has $|gH|=|H| \ \forall g\in G$ since $gH\rightarrow H$ defined by $gh\mapsto h$ is a bijection. Hence $|G|=t|H|$.
\end{proof}

\begin{defn}
\begin{enumerate}
\item As in the context of above, we write $G/H:=\{gH:g\in G\}$.
\item $|G/H|$ is called \textit{index} of $H$ in $G$, denoted $|G:H|$. By Lagrange's theorem if $G$ is finite then $|G:H|=\frac{|G|}{|H|}$.
\end{enumerate}
\end{defn}
\begin{coro}
If $G$ is finite and $g\in G$, then $|g|\mid |G|$.
\end{coro}
\begin{proof}
This follows from the fact $|\la g\ra |=|g|$ and Lagrange's theorem.
\end{proof}

\subsection{Normal subgroup and quotient group}
In general $G/H$ is not a group, which is the motivation of this section.
\begin{lemma}
Let $H\leq G,\ g\in G$. Then $gHg^{-1}=\left\{ghg^{-1}:h\in H\right\} \leq G$.
\end{lemma}
\begin{proof}
We use Lemma \ref{lemma:sgptest}. Clearly $gHg^{-1}\neq \varnothing$ since $1_G\in gHg^{-1}$. Now let $x=gh_1g^{-1},\ y=gh_2g^{-1}$ where $h_1,h_2\in H$. Note that $h_1h_2\in H$ since $H\leq G$. Then $y^{-1}=gh_2^{-1}g^{-1}$ so
\[
xy^{-1}=gh_1g^{-1}gh_2^{-1}g^{-1}=gh_1h_2^{-1}g^{-1} \in gHg^{-1}.
\]
\end{proof}
\begin{defn}
\begin{enumerate}
\item $H\leq G$ is \textit{normal} in $G$ if $gHg^{-1}=H \ \forall g\in G$, denoted $N\unlhd G$.
\item The \textit{normaliser} of $H\leq G$ is defined as
\[
N_G(H):=\{g\in G:gHg^{-1}=H\}.
\]
\end{enumerate}
\end{defn}
\begin{exe}
\begin{enumerate}
\item If $H\leq G$, show that $N_G(H)\leq G$.
\item $\{1_G\},G$ are always normal.
\end{enumerate}
\end{exe}

\begin{flushright}
\textit{Week 2, lecture 3 starts here}
\end{flushright}

\begin{defn}
$G$ is \textit{simple} if $\{1_G\}$ and $G$ are the only normal subgroups of $G$.
\end{defn}
\begin{example}
\begin{itemize}
\item $\Z/p\Z$ for any prime $p$ (by Lagrange's)
\item $A_n$ for $n\geq 5$
\end{itemize}
\end{example}
\begin{notation}
$AB:=\{ab:a\in A,b\in B\}$ where $A,B\subseteq G$. It's a subset but not a subgroup of $G$ in general, even if $A,B\leq G$.
\end{notation}
\begin{lemma}
Let $N\unlhd G$ and $g,h\in G$. Then $(gN)(hN)=ghN$.
\end{lemma}
\begin{proof}
\begin{itemize}
\item[$\subseteq$:] Let $x=gn_1\in gN,\ y=hn_2\in hN$ where $n_{1,2}\in N$. Then
\[
xy=gn_1hn_2=ghh^{-1}n_1hn_2 \in ghN
\]
since $h^{-1}n_1h\in N$ by definition of a normal subgroup.
\item[$\supseteq$:] Let $x=ghn\in ghN$ where $n\in N$. Then
\[
x=(g1_G)(hn)\in (gN)(hN).
\]
\end{itemize}
\end{proof}
\begin{defn}
Let $N\unlhd G$.
\begin{enumerate}
\item The \textit{natural binary operation} on $G/N$ is $\circ:G/N\times G/N\rightarrow G/N$ given by $(gN)\circ(hN)=ghN$.
\item $(G/N,\circ)$ is a group, called the \textit{quotient of} $G$ \textit{by} $N$.
\end{enumerate}
\end{defn}
Checking this is indeed a group is left as an exercise.

\subsection{Homomorphism}
\begin{defn}
Let $G,H$ be groups.
\begin{enumerate}
\item A map $\theta:G\rightarrow H$ is a \textit{homomorphism} if $\theta(g_1g_2)=\theta(g_1)\theta(g_2) \ \forall g_1,g_2\in G$.
\item A bijective homomorphism is an \textit{isomorphism}. If for $G,H,\ \exists\theta:G\rightarrow H$ an isomorphism, then $G$ and $H$ are \textit{isomorphic}, denoted $G\cong H$.
\item Let $\theta:G\rightarrow H$ be a homomorphism. The \textit{kernel} of $\theta$, denoted $\ker \theta$, is defined to be $\{g\in G:\theta (g)=1_H\}$, which is a subgroup of $G$. The \textit{image} of $\theta$, denoted $\im \theta$, is defined to be $\{\theta(g):g\in G\}$.
\end{enumerate}
\end{defn}
\begin{example}
Let $F$ be a field, $G=GL_n(F)$ and $H=F^\times$. Then $\det G\rightarrow H$ is a (surjective) homomorphism, since $\det AB=\det A\det B \ \forall A,B\in GL_n(F)$. Also
\[
\ker \det = \{A\in GL_n(F):\det A=1_F\}=: SL_n(F).
\]
\end{example}

\begin{thm}[1st isomorphism theorem]
Let $\theta:G\rightarrow H$ be an homomorphism. Then
\begin{enumerate}
\item $\ker\theta\unlhd G$.
\item $\im \theta\leq H$.
\item $G/\ker\theta\cong\im\theta$.
\end{enumerate}
\end{thm}

\begin{thm}[2nd isomorphism theorem]
\label{thm:2ndiso}
Let $H\leq G$ and $N\unlhd G$. Then
\begin{enumerate}
\item $HN=NH\leq G$.
\item $H\cap N\unlhd H$.
\item $HN/N\cong H/(H\cap N)$.
\end{enumerate}
\end{thm}

\begin{thm}[3rd isomorphism theorem]
\label{thm:3rdiso}
Let $N,K\unlhd G : N\leq K$. Then
\[
K/N \unlhd G/N \quad \text{and} \quad (G/N)/(K/N)\cong G/K.
\]
\end{thm}

\begin{thm}[Correspondence (or 4th isomorphism) theorem]
\label{thm:correspondence}
Let $N\unlhd G$. Then the map
\[
f:\{J:N\leq J\leq G\} \rightarrow \{X:X\leq G/N\}
\]
given by
\[
J\mapsto J/N
\]
is a bijection.
\end{thm}
\begin{proof}
Let $A:=\{J:N\leq J\leq G\}$ and $B:=\{X:X\leq G/N\}$. Clearly $J/N\leq G/N$.

Suppose $J_{1,2}\in A$ and $f(J_1)=f(J_2)$, and let $x\in J_1$. Then
\[
xN\in f(J_1)=f(J_2)=J_2/N,
\]
so $xN=yN$ for some $y\in J_2$. Since $x\in xN$, $x=yn\in J_2$ for some $n\in N$. It follows that $J_1\subseteq J_2$, and symmetrically $J_2\subseteq J_1$. Hence $f$ is injective.

Let $X\in B$ and set $Y=\{y\in G:yN\in X\}$. One can see that $Y\leq G$ since $y_{1,2}N\in X\Rightarrow (y_1N)(y_2N)^{-1}\in X\Rightarrow y_1y_2^{-1}N \in X$, so $y_1y_2^{-1}\in Y$ by definition, hence $Y\leq G$. Since $N\leq Y$ ($nN=N=1_{G/N}\in X \ \forall n\in N$) one has $y\in A$. Since $f(Y)=X,\ f$ is surjective.
\end{proof}

\begin{flushright}
\textit{Week 3, lecture 1 starts here}
\end{flushright}

\section{Group action}
\subsection{Permutation group}
\begin{defn}
Let $X$ be a set. $G\leq\Sym(X)$ is called a \textit{permutation group} on $X$.
\end{defn}

\begin{defn}
\begin{enumerate}
\item Let $g\in\Sym(X)$. The \textit{support} of $g$ is defined
\[
\supp(g):=\{x\in X:g(x)\neq x\} \subseteq X.
\]
\item Let $G\leq \Sym(X)$. The \textit{support} of $G$ is defined
\[
\supp(G):=\{x\in X:g(x)\neq x\text{ for some }g\in G\} \subseteq X.
\]
\end{enumerate}
\end{defn}
\begin{example}
\begin{enumerate}
\item $\supp(\Sym(X))=X$.
\item $\supp(\{1_G\})=\varnothing$.
\item $X=\{1,2,3,4,5,6\}$ and $g=(1,5,6)$. Then $\supp(g)=\{1,5,6\}$.
\item $X=\{1,2,3,4,5\}$ and $g=(1,2)(3,5)$. Then $\supp(g)=\{1,2,3,5\}$.
\end{enumerate}
\end{example}
\begin{remark}
As the above examples show, one can read off the support of $g\in\Sym(X)$ from its decomposition as a product of disjoint cycles. More precisely, if $f\in\Sym(X)$, $f=f_1\ldots f_m$ is such decomposition where $f_i=\left(a_{i_1},\ldots,a_{i_{t_i}}\right)$. Then
\[
\supp(f)=\{a_{i_j} : 1\leq i\leq m, \ 1\leq j \leq t_i\}.
\]
\end{remark}
\begin{exe}
Let $H,G\leq \Sym(X)$.
\begin{enumerate}
\item Show that $H\leq G\Rightarrow\supp(H)\subseteq\supp(G)$.
\item Deduce that $\supp(H)\cap\supp(G)\Rightarrow H\cap G =\{1_{\Sym(X)}\}$.
\item Is the converse of above true?

No, counterexample: $X=\{1,2,3\},\ G=\la(1,2)\ra,\ H=\la(2,3)\ra$.
\item What if $gh=hg \ \forall g\in G,h\in H$?
\end{enumerate}
\end{exe}

\begin{thm}
\label{thm:disjointcycle}
\begin{enumerate}
\item Disjoint cycles commute.
\item Let $f\in\Sym(X)$ and $f=f_1\ldots f_m$ as a product of disjoint cycles $f_i$. If $m=1$ then $|f|$ is length of $f_1$. If $m\geq 2$ then $|f|=\lcm (|f_1|,\ldots,|f_m|)$.
\item If $f=(a_1,\ldots,f_r)\in\Sym(X)$ is a cycle and $g\in\Sym(X)$, then $^gf:=gfg^{-1}=\left(g(a_1),\ldots,g(a_r)\right)$.
\end{enumerate}
\end{thm}

\begin{proof}[Proof (nonexaminable)]
\begin{enumerate}
\item Let $f=(a_1,\ldots,a_r),\ g=(b_1,\ldots,b_s)$ be disjoint cycles. One needs to prove $(f\circ g)(x)=(g\circ f)(x) \ \forall x\in X$.

Suppose $x\in\{a_1,\ldots,a_r\}$, which implies $x\neq b_i$ by assumption. So $g(x)=x$ by definition of cycles, hence $f(g(x))=f(x)$. Also, again by definition, $f(x)\in\{a_1,\ldots,a_r\}$, so $f(x)\neq b_i$, hence $g(f(x))=f(x)$. The argument for case $x\notin \{a_1,\ldots,a_r\}$ is symmetric.
\item The case $m=1$ is seen before in section 1.3. We prove the claim by induction on $m$. Suppose $m\geq 2$ and all precedents are true. Let $g=f_1\ldots f_{m-1}$. We now need three things to finish the proof:
\begin{enumerate}
\item Write $f_i=(a_{i_1},\ldots,a_{i_{t_i}})$. Then $\supp(g)=\{a_{i_j}:1\leq i\leq m-1,\ 1\leq j\leq t_i\}$ and $\supp(f_m)=\{a_{m_j}:1\leq j\leq t_m\}$. By assumption $\supp(g)\cap\supp(f_m)=\varnothing$, so $\la g\ra \cap \la f_m\ra=\{1_{\Sym(X)}\}$ by exercise above.
\item $g$ and $f_m$ commute by 1.
\item $|g|=\lcm(|f_1|,\ldots,|f_{m-1}|)$ by inductive hypothesis.
\end{enumerate}
By Lemma \ref{lemma:orderlem}.4 one has the desired.
\item Let $b_i:=g(a_i)$ and observe that $(gfg^{-1})(b_i)=gfg^{-1}(g(a_i))=g(f(a_i))=g(a_{i+1})=b_{i+1}$. Now let $x\in X\backslash\{b_1,\ldots,b_m\}$. Then $g^{-1}(x)\in X\backslash\{g^{-1}(b_1),\ldots,g^{-1}(b_m)\}$ since $g$ is a bijection, i.e. $g^{-1}(x)\in X\backslash\{a_1,\ldots,a_m\}$, so $f(g^{-1}(x))=g^{-1}(x)$, and $gfg^{-1}(x)=g(g^{-1}(x))=x$.
\end{enumerate}
\end{proof}

\begin{flushright}
\textit{Week 3, lecture 2 starts here}
\end{flushright}
Recall that a subgroup of $G$ generated by a nonempty $A\subseteq G$ is defined to be
\[
\la A\ra :=\left\{ a_1^{\varepsilon_1}\cdots a_m^{\varepsilon_m}:m\in\N,\ \varepsilon_i\in\{\pm 1\},\ a_i\in A \right\}.
\]
\begin{exe}
\label{exe:checkgeneratingset}
Let $A\subseteq G$ be nonempty.
\begin{enumerate}
\item Show that
\[
\la A\ra = \bigcap_{A\subseteq H\leq G} H.
\]
In particular, if $H\leq G$ and $A\subseteq H$ then $\la A\ra\leq H$.
\item Recall that given $H\leq G,\ N_G(H):=\{g\in G:gHg^{-1}=H\}$. Suppose $g\in G$ and $gag^{-1}\in\la A\ra \ \forall a\in A$. Show that $g\in N_G(\la A\ra).$ (One only needs to check element in generating set instead of the whole subgroup for normaliser.)
\end{enumerate}
\end{exe}
\begin{defn}
\label{defn:dihedralgp}
Let $n\in\N,\ n\geq 3$ and set $X:=\{1,\ldots,n\}$. Define $\sigma,\tau\in\Sym(X)$ by $\sigma:=(1,2,\ldots,n)$ and $\tau=\prod_{i=1}^{\lfloor n/2\rfloor} (i,n-i+1)=(1,n)(2,n-1)\cdots$. The \textit{dihedral group of order} $2n$ is the permutation group on $X$ defined by $D_{2n}:=\la\sigma,\tau\ra.$
\end{defn}
This is the rigorous (algebraic) definition of $D_{2n}$, but it can also be thought of group of symmetries of a regular $n$-gon.
\begin{example}
\begin{enumerate}
\item $n=8,\ \sigma=(1,2,3,4,5,6,7,8),\ \tau=(1,8)(2,7)(3,6)(4,5)$.
\item $n=7,\ \sigma=(1,2,3,4,5,6,7),\ \tau=(1,7)(2,6)(3,5)$.
\end{enumerate}
\end{example}

\begin{thm}
Let $n\in\N,\ n\geq 3$.
\begin{enumerate}
\item $|D_{2n}|=2n$.
\item $N:=\la\sigma\ra \unlhd D_{2n}$ and $|N|=n$.
\end{enumerate}
\end{thm}
\begin{proof}
\begin{enumerate}
\item See sheet 2.
\item First note that $\tau\sigma\tau^{-1}=(\tau(1),\ldots,\tau(n))=(n,n-1,\ldots,1)=\sigma^{-1}$ by Theorem \ref{thm:disjointcycle}.3 and definition of $\tau$. Also clearly $\sigma\sigma\sigma^{-1}=\sigma$. Now if $A:=\{\sigma\}$ then we have shown $^\tau \sigma,^\sigma \sigma \in\la A\ra$, so by Exercise \ref{exe:checkgeneratingset}.2, $\tau,\sigma\in N_{D_{2n}}(\la A\ra)$. Hence $\la\{\tau,\sigma\}\ra=D_{2n}\subseteq N_{D_{2n}}(\la A\ra)$, i.e. $\la A\ra \unlhd D_{2n}$. Also $|N|=|\la \sigma\ra|=|\sigma|=n$.
\end{enumerate}
\end{proof}

\begin{defn}
Let $X$ be a finite set.
\begin{enumerate}
\item Let $f\in\Sym(X)$ and write $f=f_1\cdots f_m$ as product of disjoint cycles. $f$ is \textit{even} if the number of cycles of even length in $\{f_1,\ldots,f_m\}$ is even. Otherwise $f$ is \textit{odd}.
\item The \textit{alternating group on} $X$, denoted $\Alt(X)$, is defined $\{f:f\in\Sym(X)\text{ even}\}$.
\end{enumerate}
\end{defn}
\begin{example}
$(1,2,3,4)\in S_4$ is odd, $(1,2)(3,4,5)\in S_5$ is odd, $(1,2)(3,4,5,6)\in S_6$ is even.
\end{example}

\begin{prop}
$\Alt(X)\leq \Sym(X)$ and $[\Sym(X):\Alt(X)]=2$, i.e. $|\Alt(X)|=\frac{|X|!}{2}$.
\end{prop}
\begin{proof}
See sheet 2.
\end{proof}

\begin{prop}
If $X,Y$ are finite sets with $|X|=|Y|$, then $\Sym(X)\cong\Sym(Y)$.
\end{prop}
\begin{proof}
Let $\beta:X\rightarrow Y$ be a bijection. Define $\theta:\Sym(X)\rightarrow\Sym(Y)$ by $f\mapsto \beta f\beta^{-1}$. It's then clear that $\theta$ is an isomorphism.
\end{proof}

\begin{flushright}
\textit{Week 3, lecture 3 starts here}
\end{flushright}

Recall that if $G=\la B\ra,\ H=\la A\ra$, then $H\unlhd G\Leftrightarrow bab^{-1}\in H \ \forall a\in A, b\in B$.

\subsection{Group action}
\begin{defn}
Let $G$ be a group and $X$ a set. An \textit{action} of $G$ on $X$ is a map $\cdot:G\times X\rightarrow X$ such that
\begin{enumerate}
\item $1_G\cdot x=x \quad \forall x\in X$
\item $(gh)\cdot x = g\cdot (h\cdot x) \quad \forall g,h\in G,\ x\in X$
\end{enumerate}
We say $G$ \textit{acts on} $X$ and $X$ is a a $G$\textit{-set}.
\end{defn}

\begin{example}
\label{example:gpaction}
\begin{enumerate}
\item The action of $G$ on itself by left multiplication: let $X:=G$ and define $\cdot :G\times X\rightarrow X$ by $g\cdot x:= gx,\ g\in G,\ x\in X$. Note that by definition of a group,
\begin{enumerate}
\item $1_G\cdot x=1_Gx=x \quad \forall x\in X$,
\item $(gh)\cdot x = (gh)x=g(hx)=g\cdot(h\cdot x) \quad \forall g,h\in G,\ x\in X$.
\end{enumerate}
\item The action of $G$ on itself by conjugation: again let $X:=G$. Define $\cdot :G\times X\rightarrow X$ by $g\cdot x := gxg^{-1}$. Note that
\begin{enumerate}
\item $1_G\cdot x = 1_G x 1_G^{-1}=x \quad \forall x\in X$,
\item $(gh)\cdot x=(gh)x(gh)^{-1}=ghxh^{-1}g^{-1}=g\cdot \left(hxh^{-1}\right)=g\cdot (h\cdot x)$.
\end{enumerate}
\item The action of $G$ on the set of left cosets of $H\leq G$: let $X:=G/H=\{gH:g\in G\}$ and define $\cdot : G\times X\rightarrow X$ by $g\cdot kH=gkH$. To see it's indeed an action is similar to 1.
\end{enumerate} 
\end{example}

\begin{prop}
\label{prop:GhomSymX}
Let $G$ be a group acting on a set $X$. Define $\phi:G\rightarrow\Sym(X)$ by $\phi(g)(x):=g\cdot x$. Then $\phi$ is a homomorphism. (Then $G/\ker\phi\cong H$ where $H\leq \Sym(X)$).
\end{prop}
\begin{proof}
Let $g,h\in G$. $\phi$ is indeed a bijection by definition of an action. It suffices to show $\phi(gh)=\phi(g)\circ\phi(h)$. Let $x\in X$, then
\[
\phi(gh)(x)=(gh)\cdot x = g\cdot(h\cdot x)=\phi(g)(\phi(h)(x))=(\phi(g)\circ\phi(h))x.
\]
\end{proof}

\begin{defn}
Let $\phi$ be the same map as above.
\begin{enumerate}
\item The \textit{kernel of action} of $G$ on $X$, denoted $\ker(G,X,\cdot)$, is defined to be
\[
\ker(G,X,\cdot)=\ker\phi=\{g\in G:g\cdot x=x \ \forall x\in X\} \unlhd G.
\]
\item The \textit{image} of the action, denoted $\im(G,X,\cdot)$, is defined to be $\im\phi\leq\Sym(X)$.
\item The action is \textit{trivial} if $\ker(G,X,\cdot)=G$ and \textit{faithful} if $\ker(G,X,\cdot)=\{1_G\}$.
\end{enumerate}
\end{defn}

\begin{example}[The same ones from \ref{example:gpaction}]
\begin{enumerate}
\item $\ker(G,X,\cdot)=\{1_G\}$, a faithful action.
\item $\ker(G,X,\cdot)=\{g\in G:gxg^{-1}=x \ \forall x\in X\}=Z(G)$. The action is trivial iff $G$ is abelian.
\item Observe that the action is trivial $\Leftrightarrow gxH=xH \ \forall g,x\in G\Leftrightarrow H=G$, i.e. it's nontrivial as long as $H$ is proper. This is useful: let $G$ be a nonabelian finite simple group. We claim $G$ cannot have a subgroup of index 3 (the case that index is 2 is obvious since if that's true then it has a nontrivial proper normal subgroup, so not simple).
\begin{proof}
Suppose $|G:H|=3.\ G$ acts on $X:=G/H$ and by the above $H$ is proper, so $K:=\ker(G,X,\cdot) \unlhd G$ is proper. But $G$ is simple so $K=\{1_G\}$ and one can then say $G\cong G/K\cong$ some subgroup of $S_3$. Since it's nonabelian it must be the whole group. But $S_3$ is not simple, a contradiction. 
\end{proof}
\end{enumerate}
\end{example}

\begin{flushright}
\textit{Week 4, lecture 1 starts here}
\end{flushright}

\begin{remark}
We saw last time that Proposition \ref{prop:GhomSymX} is particularly useful when $G$ is a finite simple group and $H$ is a subgroup of $G$ such that $|G:H|=n$, in that it implies that $G$ is isomorphic to a subgroup of $S_n$. This leads to the following more general result.
\end{remark}

\begin{prop}
\label{prop:faithfulisosgpsymX}
Let $G$ be a group acting faithfully on a set $X$. Then $G$ is isomorphic to a subgroup of $\Sym(X)$.
\end{prop}
\begin{proof}
This follows immediately from the definition of faithful and the 1st isomorphism theorem.
\end{proof}

\begin{defn}
Let $G$ be a group acting on a set $X$ and $x\in X$.
\begin{enumerate}
\item The \textit{orbit} of $x$ is $\orb_G(x):=\{g\cdot x:g\in G\}$.
\item The \textit{stabiliser} of $x$ is $\stab_G(x):=\{g\in G:g\cdot x=x\}$.
\end{enumerate}
\end{defn}

\begin{prop}
\begin{enumerate}
\item $\stab_G(x)\leq G$.
\item $\displaystyle \ker (G,X,\cdot)=\bigcap_{x\in X} \stab_G(x)$.
\end{enumerate}
\end{prop}
\begin{proof}
See sheet 2 Q8.
\end{proof}

\begin{example}[From \ref{example:gpaction}.2]
Fix $x\in X=G$. One has
\[
\orb_G(x)=\{gxg^{-1}:g\in G\},
\]
called the \textit{conjugacy class} of $x$ in $G$, sometimes denoted $^G x$. Also
\[
\stab_G(x)=\{g\in G:gxg^{-1}=x\},
\]
called the \textit{centraliser} of $x$ in $G$, sometimes denoted $C_G(x)$.
\end{example}

\begin{thm}[Orbit–stabiliser]
Let $G$ be a finite group acting on a set $X$ and $x\in X$. Then
\[
|G:\stab_G(x)|=|\orb_G(x)|,
\]
or alternatively
\[
|G|=|\stab_G(x)||\orb_G(x)|.
\]
\end{thm}
\begin{proof}
Let $S=\stab_G(x)$. Recall $G/S=\{gS:g\in G\}$ and $|G:S|=|G/S|$. Define
\[
f:G/S\rightarrow\orb_G(x) \text{ by } gS\mapsto g\cdot x.
\]
It suffices to show $f$ is bijective.
\begin{enumerate}
\item $f$ is well-defined and injective: $gS=kS\Leftrightarrow k^{-1}g\in S \Leftrightarrow k^{-1}g\cdot x=x \Leftrightarrow g\cdot x=k\cdot x\Leftrightarrow f(gS)=f(kS)$;
\item For $g\cdot x\in\orb_G(x)$ then $f(gS)=g\cdot x$, so $f$ is surjective.
\end{enumerate}
\end{proof}

\begin{coro}
\label{coro:orbpartitionX}
\begin{enumerate}
\item For $x,y\in X$, either $\orb_G(x)=\orb_G(y)$ or $\orb_G(x)\cap \orb_G(y)=\varnothing$.
\item $\{\orb_G(x):x\in X\}$ is a partition of $X$.
\item $|\orb_G(x)|$ divides $|G|$.
\end{enumerate}
\end{coro}
\begin{proof}
\begin{enumerate}
\item[1, 2.] Define a relation on $X$ $x\sim y$ if $y=g\cdot x$. It follows from the definition of an action that $\sim$ is an equivalence relation and the equivalence classes are $\{\orb_G(x):x\in X\}$.
\item[3.] Immediate from the theorem.
\end{enumerate}
\end{proof}

\begin{thm}[Cayley's]
Let $G$ be a finite group. Then $G$ is isomorphic to a subgroup of $\Sym(X)$ for some set $X$.
\end{thm}
\begin{proof}
By Example \ref{example:gpaction}.1, $G$ acts on itself by left multiplication, and $\ker(G,X,\cdot)=\{1_G\}$, i.e. the action is faithful. The result then follows from Proposition \ref{prop:faithfulisosgpsymX}.
\end{proof}

\begin{thm}
\label{thm:primepowerordergpnontrivZ}
Let $p$ be prime and $G$ a group of order $p^n$ where $n\in\N^+$. Then $|Z(G)|>1$.
\end{thm}
\begin{proof}
Observe that
\[
g\in Z(G)\Leftrightarrow gxg^{-1}=x \ \forall x\in G \Leftrightarrow xgx^{-1}=g \Leftrightarrow |\orb_G(g)|=1.
\]

\begin{flushright}
\textit{Week 4, lecture 2 starts here}
\end{flushright}

Let $\orb_G(x_1),\ldots,\orb_G(x_t)$ be the orbits of $G$ in its action by conjugation on $X=G$ (Example \ref{example:gpaction}.2). Assume WLOG that $|\orb_G(x_i)|=1$ for $1\leq i\leq s$ and $|\orb_G(x_i)|>1$ for $s<i\leq t$. By the observation above, one then has $Z(G)=\{x_1,\ldots,x_s\}$ and in particular, $|Z(G)|=s$. If $s<i\leq t$, then $|\orb_G(x)|=p^{a_i}$ for some $a_i\in \N$ by Corollary \ref{coro:orbpartitionX}.3. Now, by Corollary \ref{coro:orbpartitionX}.2,
\[
|G|=|X|=\sum_{i=1}^t \left|\orb_G(x_i)\right|=s+\sum_{i=s+1}^t p^{a_i}=p^n,
\]
so $|Z(G)|=s\equiv 0\Mod p$, hence $|Z(G)|\neq 1$.
\end{proof}

\begin{remark}
Many groups we shall see in the course will have a trivial centre, e.g. $S_n$ for $n\geq 3$ and $D_{2n}$ for $n\geq 3$. Also, a nonabelian finite simple group is not of order $p^n$.
\end{remark}

\begin{coro}
\label{coro:orderp2gpabelian}
Let $p$ be prime and $G$ a group.
\begin{enumerate}
\item $|G|=p^2\Rightarrow G$ is abelian.
\item $|G|=p^3\Rightarrow$ either $G$ is abelian or $|Z(G)|=p$.
\end{enumerate}
\end{coro}
\begin{proof}
We need two facts:
\begin{enumerate}
\item All groups of order $p$ are cyclic (immediate from Lagrange).
\item If $G$ is nonabelian then $G/Z(G)$ is not cyclic (see sheet 2 Q1).
\end{enumerate}
It follows that if $G$ is nonabelian then $|G/Z(G)|\neq p$ for a prime $p$. Now Theorem \ref{thm:primepowerordergpnontrivZ} implies
\begin{enumerate}
\item $|G|=p^2\Rightarrow |Z(G)|=p^2\Rightarrow Z(G)=G \Rightarrow G$ is abelian.
\item $|G|=p^3\Rightarrow |Z(G)|=p$ or $p^3$ and the desired result is clear.
\end{enumerate}
\end{proof}

\begin{thm}[Cauchy's]
Let $G$ be a finite group and $p$ a prime divisor of $|G|$. Then $G$ has an element of order $p$. Furthermore, number of elements of order $p$ is congruent to $-1\Mod p$.
\end{thm}
\begin{proof}
Define
\[
X:=\{(g_1,\ldots,g_p)\in G^p:g_1\cdots g_r=1_G\}.
\]
Note that
\[
\begin{aligned}
x=(g_1,\ldots,g_p)\in X &\Rightarrow 1_G=g_1\cdots g_p \\
&\Rightarrow g_i^{-1}\cdots g_1^{-1} 1_G g_1\cdots g_i=g_i^{-1}\cdots g_1^{-1} g_1\cdots g_p g_1\cdots g_i \\
&\Rightarrow 1_G=g_{i+1}\cdots g_pg_1\cdots g_i \\
&\Rightarrow (g_{i+1},\ldots ,g_p,g_1,\ldots ,g_i)\in X.
\end{aligned}
\]
Now define
\[
C:=\la\sigma\ra\leq S_p \text{ where }\sigma=(1,2,\ldots,p)
\]
and the action
\[
\cdot :C\times X\rightarrow X \text{ by } \sigma^i\cdot (g_1,\ldots,g_p):=(g_{i+1},\ldots ,g_p,g_1,\ldots ,g_i).
\]
(Check $\cdot$ is indeed an action.) Now
\begin{enumerate}
\item If $g\in G$ and $g^p=1_G$ then $(g,\ldots,g)\in X$, and $\sigma^i\cdot(g,\ldots,g)=(g,\ldots,g) \ \forall i$, i.e. $|\orb_C((g,\ldots,g))|=1$.
\item We claim that the converse is true: if $x$ satisfies $|\orb_C(x)|=1$ then $x=(g,\ldots,g)$ for some $g\in G : g^p=1_G$. Indeed, say $x=(g_1,\ldots,g_p)$. It suffices to show $g_1=g_i \ \forall i$. By the Orbit–stabiliser theorem, $|\orb_C(x)|=1$ implies $\stab_C(x)=C$, i.e. $\forall i,$
\[
(g_1,\ldots,g_p)=\sigma^{i-1}(g_1,\ldots,g_p)=(g_i,\ldots ,g_p,g_1,\ldots ,g_{i-1}),
\]
which gives the desired.
\item Note that if $(g_1,\ldots,g_p)\in X$ then $g_p=(g_1\cdots g_{p-1})^{-1}$. We claim $|X|=|G|^{p-1}$. Indeed, define $f:X\rightarrow G^{p-1}$ by $(g_1,\ldots,g_p)\mapsto (g_1,\ldots,g_{p-1})$. It suffices to show that $f$ is bijective since then $|X|=|G^{p-1}|=|G|^{p-1}$. To see $f$ is injective, note that
\[
\begin{aligned}
f((g_1,\ldots,g_p))=f((h_1,\ldots,h_p)) &\Rightarrow g_i=h_i \text{ for } 1\leq i\leq p-1 \\
&\Rightarrow g_p=(g_1\cdots g_{p-1})^{-1}=(h_1\cdots h_{p-1})^{-1}=h_p \\
&\Rightarrow (g_1,\ldots,g_p)=(h_1,\ldots,h_p).
\end{aligned}
\]
To see $f$ is surjective, note that for every $(x_1,\ldots,x_{p-1})\in G^{p-1}$ one can set $x_p:=(x_1\cdots x_{p-1})^{-1}$, then $(x_1,\ldots,x_p)\in X$ and it satisfies $f((x_1,\ldots,x_p))=(x_1,\ldots,x_{p-1})$.
\end{enumerate}
By Corollary \ref{coro:orbpartitionX}.3, all orbits not of size 1 have size $p$. Let $s$ be number of distinct orbits of size 1, $t$ be number of distinct orbits of size $p$ and $r$ be number of elements of order $p$ in $G$. By parts 1 and 2, $s=1+r$ where 1 corresponds to the trivial element $(1_G,\ldots,1_G)$. One can then write $|G|^{p-1}=|X|=1+r+pt$, and since $p\mid |G|,\ r\equiv -1\Mod p$. In particular, $r>0$.
\end{proof}

\begin{flushright}
\textit{Week 4, lecture 3 starts here}
\end{flushright}

\begin{tool}[Analysing element orders in a finite group]
\label{tool:elementorder}
Let $E_p(G):=\{x\in G:|x|=p\}$ where $p$ prime. Then
\begin{enumerate}
\item $|E_p(G)|\equiv -1\Mod p$ (Cauchy's theorem)
\item $|E_p(G)|\geq |G:C_G(x)| \ \forall x\in G$ by \ref{exe:conjaugatesordersequal}.1 and the Orbit–stabiliser theorem.
\item If $r\neq p$ is a prime and $G$ has no element of order $pr$, then $|C_G(x)|$ is not divisible by $r$ for $x\in E_p(G)$ by Lemma \ref{lemma:orderlem}.4 and Cauchy's theorem.
\end{enumerate}
\end{tool}

\begin{example}
Let $G$ be of order 48 with no elements of order 6. We claim $|E_3(G)|\geq 17$.
\begin{proof}
Let $x\in E_3(G)$. Tool \ref{tool:elementorder}.3 implies $|C_G(x)|$ is not divisible by 2. Since $|C_G(x)|\mid 48$, it must be $|C_G(x)|=3$. Then by Tool \ref{tool:elementorder}.2 $|E_3(G)|\geq 16$, and since $|E_3(G)|\equiv -1\Mod 3,\ |E_3(G)|\geq 17$.
\end{proof}
\end{example}

\begin{prop}
\label{prop:orderofHK}
Let $G,H,X$ be as in Example \ref{example:gpaction}.3 and $K\leq G$. Then $|KH|=\frac{|K||H|}{|K\cap H|}$.
\end{prop}
\begin{proof}
Since $G$ acts on $X$ and $K\leq G, K$ acts on $X$ as well. Let $x=H\in X$. Then
\[
\stab_K(x)=\{k\in K:kH=H\}=\{k\in K:k\in H\}=K\cap H,
\]
and
\[
|K:K\cap H|=|\orb_K(x)|=|\{kH:k\in K\}|.
\]
On the other hand,
\[
|KH|=\left|\bigcap_{k\in K} kH\right| = |\{kH:k\in K\}||H|=|K:K\cap H||H|.
\]
\end{proof}

\begin{coro}
Let $G,H,K$ as above. Then
\[
|G:H\cap K|\leq |G:H||G:K|.
\]
\end{coro}
\begin{proof}
\[
\frac{|H||K|}{|H\cap K|}=|KH|\leq |G|=\frac{|G|^2}{|G|},
\]
and rearranging gives the desired.
\end{proof}

\subsection{Fixed point}
\begin{defn}
Let $G$ be a group acting on a set $X$ and $g\in G$.
\begin{enumerate}
\item An element $x\in X$ is a \textit{fixed point} of $g$ if $g\cdot x=x$. The set of fixed points of $g$ is denoted $\fix_X(g):=\{x\in X:g\cdot x=x\}$.
\item $g$ is \textit{fixed point free} if $\fix_X(g)=\varnothing$.
\end{enumerate}
\end{defn}

\begin{lemma}[not Burnside's\footnote{William Burnside (1852–1927) was known as a pioneer in the systematic study of finite groups and indeed stated and proved this lemma, but later people found out this equality was known in as early as 1845 to Cauchy, so it's \textit{a lemma that is not Burnside's}.}]
Let $G$ be a finite group acting a finite set $X$. Then
\[
|\{\orb_G(x):x\in X\}|=:r=\frac{1}{|G|}\sum_{g\in G}|\fix_X(g)|.
\]
Informally, the number of orbits = the average number of fixed points.
\end{lemma}
\begin{proof}
We will use Corollary \ref{coro:orbpartitionX}.1 and 2. Let
\[
\Lambda = \{(g,x):g\in G,x\in X, g\cdot x=x\}.
\]
We count $|\Lambda|$ in two different ways (double-counting method to show equality).
\begin{enumerate}
\item \[
|\Lambda|=\sum_{g\in G} |\fix_X(g)|.
\]
\item \[
\begin{aligned}
|\Lambda|&=\sum_{x\in X} |\{g\in G:g\cdot x=x\}|=\sum_{x\in X} |\stab_G(x)|=\sum_{x\in X} \frac{|G|}{|\orb_G(x)|}\\
&=\sum_{i=1}^r\sum_{y\in\orb_G(x_i)} \frac{|G|}{|\orb_G(y)|}=\sum_{i=1}^r\sum_{y\in\orb_G(x_i)} \frac{|G|}{|\orb_G(x_i)|} \\
&=\sum_{i=1}^r |\orb_G(x_i)| \frac{|G|}{|\orb_G(x_i)|}=r|G|
\end{aligned}
\]
where $\orb_G(x_1),\ldots,\orb_G(x_r)$ are distinct orbits.
\end{enumerate}
\end{proof}

\begin{coro}
Let $G,X$ and $r$ be as in above lemma. Suppose $|X|>1$ and $r=1$. Then $G$ has a fixed point free element.
\end{coro}
\begin{proof}
By definition one has $|\fix_X(1_G)|=|X|$. Now
\[
1=\frac{1}{|G|}\sum_{g\in G} |\fix_X(g)|=\frac{1}{|G|}\left(|\fix_X(1_G)|+\sum_{g\neq 1_G} |\fix_X(g)|\right).
\]
So if $G$ doesn't have any fixed point free element then $|\fix_X(g)|\geq 1 \ \forall g\in G$ and
\[
1\geq \frac{1}{|G|}(|X|+|G|-1)>\frac{|G|}{|G|}=1,
\]
a contradiction.
\end{proof}

\begin{flushright}
\textit{Week 5, lecture 1 starts here}
\end{flushright}

\section{Sylow theorems}
\begin{remark}[Philosophy]
In chapter 1, we saw Lagrange's theorem. Question: does the converse hold? i.e., if $l\mid |G|$, does $G$ necessarily have a subgroup of order $l$?
\begin{enumerate}
\item A counterexample would be $A_4$ with $|A_4|=12$, which does not have a subgroup of order 6 (use Tool \ref{tool:elementorder}.3).
\item In general, let $G$ be a finite simple group of even order $>2$. Then $G$ has no subgroup of order $|G|/2$.
\end{enumerate}
Sylow theorems will prove that a partial converse holds by restricting $l$.
\end{remark}

\begin{notation}
For the remainder of the chapter, we fix a finite group $G$ and a prime divisor $p$ of $|G|$. Also, we write $|G|_p$ for the $p$-part of $|G|$, i.e. writing $|G|=p^n m$ where $p\nmid m$ we have $|G|_p=p^n$.
\end{notation}

\begin{defn}
Let $H\leq G$.
\begin{enumerate}
\item $H$ is a $p$\textit{-subgroup} of $G$ if $|H|$ is a power of $p.$
\item $H$ is a \textit{Sylow} $p$\textit{-subgroup} of $G$ if $|H|=|G|_p$.
\item The set of all Sylow $p$-subgroups of $G$ is denoted $\Syl_p(G)$.
\end{enumerate}
\end{defn}

\begin{example}
\begin{enumerate}
\item $G=S_4$ has order 24. Then $|G|_2=2^3,\ |G|_3=3$. One has $\la (1,2,3)\ra\in\Syl_3(G)$ and $D_8=\la (1,2,3,4),(1,4)(2,3)\ra\in\Syl_2(G)$. Also $\la(1,2)\ra$ is a 2-subgroup but not a Sylow 2-subgroup.
\item $G=C_n$. Then for each divisor $d$ or $n,\ G$ has a unique subgroup of order $d$. In particular, if $p\mid n$, then $|\Syl_p(G)|=1$. See sheet 2 Q3.
\item $G=GL_2(F)$ where $F$ is a field of order $p$. Then by Theorem \ref{thm:orderofGLnF}, $|G|=p^{\binom{2}{2}} \prod_{i=1}^2(p^i-1)=p(p-1)(p^2-1)$. One has $x=\begin{pmatrix}1&1\\0&1\end{pmatrix}\in G$ with order $p$. Hence $\la x\ra \in \Syl_p(G)$. More generally, $|GL_n(F)|_p=p^{\binom{n}{2}}$ and $U(n,F)$ (the set of upper triangular matrices with 1 on the diagonal) is a Sylow $p$-subgroup.
\end{enumerate}
\end{example}

\begin{thm}[Sylow theorems]
Let $G$ be a finite group with $p$ a prime divisor of $|G|$.
\begin{enumerate}
\item (Existence) $\Syl_p(G)\neq\varnothing$.
\item (Conjugacy) All Sylow $p$-subgroups are conjugate in $G$.
\item (Containment) Every $p$-subgroup of $G$ is contained in a Sylow $p$-subgroup.
\item (Number) $|\Syl_p(G)|\equiv 1\Mod p$.
\end{enumerate}
\end{thm}

\subsection{Wielandt's proof of Sylow theorems 1 \& 4}
\begin{lemma}
\label{lemma:1stlemSylow4}
Let $p$ be prime and $n,m\in\N^+$ with $\gcd(m,p)=1$. Then
\begin{enumerate}
\item $\displaystyle p\mid \binom{p}{i}$ for $1\leq i\leq p-1$.
\item $\displaystyle \binom{p^n m}{p^n}\equiv m\Mod p$.
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{enumerate}
\item Fix $1\leq i\leq p-1$. Then
\[
\binom{p}{i}=\frac{p!}{i!(p-i)!}=\frac{p(p-1)\cdots(p-i+1)}{i(i-1)\cdots 1}.
\]
Now let $a:=(p-1)\cdots (p-i+1),\ b=i!$. Then
\[
\binom{p}{i}=\frac{pa}{b}\Rightarrow pa=b\binom{p}{i}\Rightarrow p\mid b\binom{p}{i},
\]
but clearly $\gcd(p,b)=1$, hence $\displaystyle p\mid \binom{p}{i}$.
\item Let $F:=\Z/p\Z=\{0,1,\ldots,p-1\}$ with usual addition and multiplication modulo $p$. Consider the polynomial $(1+x)^p\in F[x]$.
\begin{flushright}
\textit{Week 5, lecture 2 starts here}
\end{flushright}
By binomial theorem,
\[
(1+x)^p=\sum_{i=1}^p \binom{p}{i}x^i=1+x^p \in F[x].
\]
Then
\[
(1+x)^{p^2}=\left((1+x)^p\right)^p=\left(1+x^p\right)^p=1+x^{p^2}.
\]
Inductively,
\[
(1+x)^{p^n}=1+x^{p^n}.
\]
Even more generally,
\[
(1+x)^{p^n m}=\left((1+x)^{p^n}\right)^m=\left(1+x^{p^n}\right)^m.
\]
Binomial theorem then gives us the equality
\[
\sum_{i=0}^{p^n m} \binom{p^n m}{i} x^i = \sum_{i=0}^{m} \binom{m}{i} x^{p^n i}.
\]
Comparing coefficients of $x^{p^n i}$ gives
\[
\binom{p^n m}{p^n i}=\binom{m}{i}
\]
and in particular for $i=1$,
\[
\binom{p^n m}{p^n}=m \in F.
\]
Translating this back to $\Z$ one has the desired.
\end{enumerate}
\end{proof}

\begin{prop}
Sylow theorem 4. In particular, Sylow theorem 1.
\end{prop}
\begin{proof}
As usual, write $|G|=p^n m$ where $p\nmid m$ and $p^n=:|G|_p$. Let $X:=\{S\subseteq G:|S|=|G|_p\}$. Define $\cdot G\times X\rightarrow X$ by $g\cdot S:=gS=\{gs:s\in S\}$. This is indeed an action: see sheet 2 Q12. Let $\orb_G(S_i)$ be $t$ distinct orbits in $X$. By Corollary \ref{coro:orbpartitionX}.2 and Lemma \ref{lemma:1stlemSylow4}.2,
\[
\binom{p^n m}{p^n}=|X|=\sum_{i=1}^t |\orb_G(s_i)| \equiv m\Mod p.
\]
This means at least one $|\orb_G(s_i)|$ is not divisible by $p$. WLOG, suppose $p\nmid |\orb_G(S_i)|$ for $1\leq i\leq r$ and $p\mid |\orb_G(S_i)|$ for $r< i\leq t$. We claim:
\begin{enumerate}
\item Fix $i=1,\ldots,r$ and denote $S_i$ by $S$ for convenience. Then $\exists x\in G:\stab_G(xS)=xS$ and in particular $xS\in\Syl_p(G)$. Indeed, let $s\in S$ and set $x=s^{-1},\ T:=xS$. We want to show $\stab_G(T)=T$. First note that $1_G=xx^{-1}=xs\in T$. Hence $g\in\stab_G(T)\Rightarrow gT\Rightarrow g=g1_G\in gT=T$, so $\stab_G(T)\subseteq T$. Also, $T\in\orb_G(S)$, so $\orb_G(T)=\orb_G(S)$. Hence
\[
p\nmid |\orb_G(T)|=\frac{|G|}{|\stab_G(T)|}=\frac{p^n m}{|\stab_G(T)|}.
\]
This implies $p^n\mid |\stab_G(T)|$ by Lagrange's theorem. But by construction, $|T|=p^n$, so it must be that $\stab_G(T)=T$.
\item $r=|\Syl_p(G)|$. Indeed, for $i=1,\ldots,r$ we can take $T_i=x_i S_i\in\orb_G(S_i)$ such that $T_i=\stab_G(T_i)$ by previous claim. Now define
\[
\begin{aligned}
f:\{\orb_G(T_1),\ldots,\orb_G(T_r)\} &\rightarrow \Syl_p(G) \\
\orb_G(T_i) &\mapsto T_i
\end{aligned}
\]
$f$ is well-defined since $\orb_G(T_i)$ are distinct by construction and $T_i\in\Syl_p(G)$ by first claim. Since $T_i$ are distinct, $f$ is injective. Now let $P\in\Syl_p(G)$. Then $P\in X$, and
\[
\stab_G(P)=\{g\in G:gP=P\}=P,
\]
so $|\orb_G(P)|=m$ which by definition is not divisible by $p$. Hence for some $i=1,\ldots,r,\ \orb_G(P)=\orb_G(T_i)$, so $P\in\orb_G(T_i)$, i.e. $P=gT_i$ for some $g\in G$. But $g=g1_G\in gT_i=P$ and since $g^{-1}\in P$, $T_i=g^{-1}P=P$. This proves $f$ is surjective, hence bijective, hence the claim.
\end{enumerate}
Therefore,
\[
rm+0=\sum_{i=1}^r |\orb_G(T_i)|+\sum_{i=r+1}^t |\orb_G(S_i)|=|X|\equiv m\Mod p
\]
and since $\gcd(m,p)=1,$ we can do cancellation and have $r\equiv 1\Mod p$.
\end{proof}

\begin{flushright}
\textit{Week 5, lecture 3 starts here}
\end{flushright}

\subsection{Proofs of Sylow theorems 2 \& 3}
\begin{remark}[Easy but useful facts]
Let $G$ be finite and $p$ a prime divisor of $|G|$. Then
\begin{enumerate}
\item $P\in\Syl_p(G),g\in G\Rightarrow gPg^{-1}\in\Syl_p(G)$.
\item If $|G|$ is a power of $p$ then $\Syl_p(G)=\{G\}$.
\item By definition, a $p$-subgroup $Q$ of $G$ is a Sylow $p$-subgroup iff $p\nmid |G:Q|$.
\end{enumerate}
\end{remark}

\begin{prop}
Let $G,p$ be as above and $P\in\Syl_p(G),\ H\leq G$. Then $\exists g\in G:H\cap gPg^{-1}\in\Syl_p(H)$.
\end{prop}
\begin{proof}
Let $X=G/P=\{gP:g\in G\}$. Then $H$ acts on $X$ by left multiplication (since $G$ does) (Example \ref{example:gpaction}.3). Consider the orbits and stabilisers. Fix $xP\in X$ where $x\in G$, then
\[
\begin{aligned}
\stab_H(xP)&=\{h\in H:hxP=xP\}=\{h\in H:x^{-1}hxP=P\}\\
&=\{h\in H:x^{-1}hx\in P\}=\{h\in H:h\in xPx^{-1}\}=H\cap xPx^{-1}.
\end{aligned}
\]
As usual, let $\orb_H(x_1P),\ldots,\orb_H(x_tP)$ be distinct orbits and write $|G|=p^n m$ where $p\nmid m$. We have
\[
p\nmid m=|X|=\sum_{i=1}^t \left|\orb_H(x_i P)\right|=\sum_{i=1}^t \left|H:\left(H\cap x_i Px_i^{-1}\right)\right|
\]
so $p\nmid \left|H:\left(H\cap x_i Px_i^{-1}\right)\right|$ for some $i$. We claim $g:=x_i$ satisfies the desired. Indeed, $H\cap gPg^{-1} \leq gPg^{-1}$, so by Lagrange's theorem it's a $p$-subgroup of $H$, hence by 3rd remark above it's a Sylow $p$-subgroup of $H$.
\end{proof}

\begin{coro}
Sylow theorems 2 and 3.
\end{coro}
\begin{proof}
\begin{enumerate}
\item[2.] Let $H,P\in\Syl_p(G)$. Then $\exists g\in G:H\cap gPg^{-1}\in \Syl_p(H)=\{H\}$ by previous proposition and the 2nd remark above. So $H=H\cap gPg^{-1}$, in particular $H\subseteq gPg^{-1}$, but by assumption $|H|=|gPg^{-1}|$ so $H=gPg^{-1}$.
\item[3.] Let $H\leq G$ be a $p$-subgroup and $P\in\Syl_p(G)$. Then by exactly the same argument as above, $H\subseteq gPg^{-1}\in\Syl_p(G)$.
\end{enumerate}
\end{proof}

\subsection{Consequences of Sylow theorems}
Recall that if $H\leq G$ then $H\leq N_G(H)=\{g\in G:gHg^{-1}=H\}$.

\begin{coro}
\label{coro:consequence1Sylow}
Let $G,p$ be as above and $P\in \Syl_p(G)$.
\begin{enumerate}
\item $|\Syl_p(G)|=|G:N_G(P)|$.
\item $\Syl_p(G)\mid |G:P|$.
\item $P\unlhd G\Leftrightarrow |\Syl_p(G)|=1$.
\end{enumerate}
\end{coro}
\begin{proof}
Let $G$ acts on $X:=\Syl_p(G)$ by conjugation (see sheet 2 Q15 that this is indeed an action).
\begin{enumerate}
\item By Sylow theorem 2, $\Syl_p(G)$ is explicitly $\{gPg^{-1}:g\in G\}$ which by definition is $\orb_G(P)$. Now $\stab_G(P)=\{g\in G:gPg^{-1}=P\}=N_G(P)$. The desired result then follows from Orbit–stabiliser theorem.
\item By Lagrange's theorem and part 1, $P\leq N_G(P)\Rightarrow |P| \mid |N_G(P)| \Rightarrow |G:N_G(P)| \mid |G:P|\Rightarrow |\Syl_p(G)| \mid |G:P|$.
\item We have $P\unlhd G\Leftrightarrow \{gPg^{-1}:g\in G\}=\{P\}\Leftrightarrow \Syl_p(G)=\{P\}\Leftrightarrow |\Syl_p(G)|=1$.
\end{enumerate}
\end{proof}

\begin{coro}
\label{coro:consequence2Sylow}
Let $G,p$ be as above and
\[
F_p(G):=\{x\in G:x\neq 1_G,\ |x|=p^n\}.
\]
Then
\begin{enumerate}
\item \[
F_p(G)=\bigcup_{P\in\Syl_p(G)} P\backslash \{1_G\}
\]
\item $|F_p(G)|\geq |G|_p-1$ with equality iff $|\Syl_p(G)|=1$ (i.e. there is a normal Sylow $p$-subgroup).
\item If $|G|_p=p$, then $|F_p(G)|=|\Syl_p(G)|(p-1)$.
\end{enumerate}
\end{coro}

\begin{flushright}
\textit{Week 6, lecture 1 starts here}
\end{flushright}

\begin{proof}
\begin{enumerate}
\item Let \[
x\in \bigcup_{P\in\Syl_p(G)} P\backslash \{1_G\}.
\]
Then $|x|=p^n$ by Lagrange's, and since $x\neq 1$ one has $x\in F_p(G)$. We haven't used Sylow yet. Now let $x\in F_p(G)$. Then $\la x\ra$ is a $p$-subgroup since its order is $|x|$, so $\la x\ra$ is contained in a Sylow $p$-subgroup. The desired is then clear.
\item[2, 3.] See sheet 3 Q10, 11 respectively.
\end{enumerate}
\end{proof}

\begin{example}[Applying \ref{coro:consequence1Sylow} and \ref{coro:consequence2Sylow}]
\label{example:strategy2}
\begin{enumerate}
\item Prove that a group of order 30 is not simple.
\begin{proof}
Suppose $|G|=30$ and $G$ is simple. Note $|G|=2\times 3\times 5$. By Corollary \ref{coro:consequence1Sylow}.2 and Sylow theorem 4, $|\Syl_5(G)|\mid 6$ and $|\Syl_5(G)|\equiv 1\Mod 5$, i.e. $|\Syl_5(G)|=1$ or 6. If it's 1 then by Corollary \ref{coro:consequence1Sylow}.3 $G$ is not simple with $P$ normal, a contradiction; so $|\Syl_5(G)|=6$. Similarly, $|\Syl_3(G)|=10$. Now Corollary \ref{coro:consequence2Sylow}.3 says $|F_5(G)|=6\times 4=24$ and $|F_3(G)|=10\times 2=20$, but we only have 30 elements. Hence $G$ must be not simple.
\end{proof}
\item Prove that a group of order 132 is not simple.
\begin{proof}
Suppose $|G|=132=11\times 2^2\times 3$ and $G$ is simple. Then similarly, $|\Syl_{11}(G)| \mid 12$ and $|\Syl_{11}(G)|\equiv 1\Mod 11$, i.e. $|\Syl_{11}(G)|=1$ or 12. But again $G$ has no normal subgroup, so $|\Syl_{11}(G)|=12$. Similarly, $|\Syl_3(G)|=4$ or 22. Again, $|F_{11}(G)|=12\times 10=120$ and $|F_3(G)|\geq 4\times 2=8$. Now,
\[
F_2(G)\subseteq G\backslash F_{11}(G) \sqcup F_3(G) \sqcup \{1_G\},
\]
so
\[
|F_2(G)|\leq 132-120-8-1=3.
\]
Corollary \ref{coro:consequence2Sylow}.2 says $|F_2(G)|\geq 2^2-1=3$, so $|F_2(G)|=3$, hence there is a normal Sylow $p$-subgroup, a contradiction with $G$ being simple.
\end{proof}
\end{enumerate}
\end{example}

\subsection{2 applications of Sylow theorems}
In this section, we'll look at a game with 2 versions.
\begin{itemize}
\item Version 1: Prove that a group $G$ of order $\ast$ is not simple. The 3 strategies are
\begin{enumerate}
\item Immediately apply Corollary \ref{coro:consequence1Sylow}.2 and Sylow theorem 4 to try to get a contradiction. We usually start with the largest $p$.

\textbf{e.g.} $\ast=20=2^2\times 5$. Then $|\Syl_5(G)|=1$, an immediate contradiction.
\item The $F_p(G)$-strategy: for each $p$ such that $|G|_p=p$, use Corollary \ref{coro:consequence2Sylow}.3 to get a lower bound on $|F_p(G)|$. Since 
\[
|G|<\sum_{p\mid |G|}|F_p(G)|,
\]
we either get an immediate contradiction or we should further use Corollary \ref{coro:consequence2Sylow}.3 to get one.

\textbf{e.g.} Example \ref{example:strategy2}.

\begin{flushright}
\textit{Week 6, lecture 2 starts here}
\end{flushright}

\item The homomorphism strategy: again begin by considering possibilities for $|\Syl_p(G)|$. Note that if we choose a $p$ such that $|G:N_G(P)|=|\Syl_p(G)|=m>1$ for $P\in\Syl_p(G)$ (Corollary \ref{coro:consequence1Sylow}), then $\ker(G,\Syl_p(G),\cdot)\subseteq \stab_G(P)=N_G(P) \subsetneqq G$ is proper. Since we assume (for contradiction) that $G$ is simple, $\ker(G,\Syl_p(G),\cdot)=\{1_G\}$ because otherwise it would be a nontrivial, proper normal subgroup. Hence by Proposition \ref{prop:faithfulisosgpsymX}, $G\cong$ some subgroup of $\Sym(X)$ and in particular $|G|\mid m!$. We would then get a contradiction hopefully.

\textbf{e.g.} $\ast=48=2^4\times 3$. Then $|\Syl_2(G)|=3$. So $G\cong$ a subgroup of $(\Sym(\Syl_2(G))\cong S_3)$ and in particular $48\mid 6$, which is absurd.
\end{enumerate}

\item Version 2: Prove that a finite group $G$ with given properties (usually conjugacy classes of elements of prime order) is simple. Essentially, use the following corollary.
\end{itemize}

\begin{coro}
\label{pactionhappeninN}
Let $N\unlhd G$ a finite group and $p$ a prime divisor of $|G|$. Then
\begin{enumerate}
\item $x\in N\Rightarrow \{gxg^{-1}:g\in G\}\subseteq N$.
\item $p\nmid |G:N| \Rightarrow \Syl_p(N)=\Syl_p(G)$ and $F_p(N)=F_p(G)$.
\end{enumerate}
\end{coro}
\begin{proof}
\begin{enumerate}
\item Immediate from definition.
\item By the 2nd isomorphism theorem, for a $P\in\Syl_p(G),\ P/(P\cap N)\cong PN/N\leq G/N$. So $|PN/N|\mid |P|$, hence by Lagrange's, $PN/P$ is a $p$-subgroup of $G/N$. But $p\nmid |G:N|$, so $PN/N=\{1_{G/N}\}$, i.e. $PN=N$, so $P\leq N$. So $|N|_p=|G|_p$, hence $\Syl_p(G)\subseteq \Syl_p(N)$. The other inclusion is clear.

Now $F_p(G)=\bigcup_{P\in\Syl_p(G)} P\backslash \{1_G\}=\bigcup_{P\in\Syl_p(N)} P\backslash \{1_G\}=F_p(N)$.
\end{enumerate}
\end{proof}

\begin{thm}
$A_5$ is simple.
\end{thm}
\begin{proof}
We need 4 facts about $G=A_5$ to start with:
\begin{enumerate}
\item $|G|=60=2^2\times 3\times 5$.
\item $G$ has 24 elements of order 5, the 5-cycles.
\item $G$ has 20 elements of order 3, the 3-cycles.
\item $G$ has 15 elements of order 2, precisely of the form $(a,b)(c,d)$ where $a,b,c,d \in \{1,\ldots,5\}$ are distinct and all such elements are conjugate.
\end{enumerate}

\begin{flushright}
\textit{Week 6, lecture 3 starts here}
\end{flushright}

Suppose $G$ is not simple and let $N\unlhd G$.
\begin{enumerate}
\item[$1^\circ$:] $p\mid |N|$ for some $p\in\{3,5\}$. Then since $|G|_p=p,\ p\nmid |G:N|$. So $F_p(G)=F_p(N)$. Hence
\begin{itemize}
\item $p=5\Rightarrow |N|\geq |F_5(N)|+1\geq 25$
\item $p=3\Rightarrow |N|\geq |F_3(N)|+1\geq 21$
\end{itemize}
so Lagrange's implies $|N|=30$, i.e. both 3 and 5 divide $|N|$. But again by Corollary \ref{pactionhappeninN}
\[
|N|\geq |F_3(N)|+|F_5(N)|+1\geq 45,
\]
a contradiction.
\item[$2^\circ$:] Neither 3 nor 5 divides $|N|$, then $|N|\mid 4$, so by Cauchy's it contains an element of order 2. Hence $N$ contains all elements of order 2, so $15\leq |N|\mid 4$, a contradiction.
\end{enumerate}
\end{proof}

\begin{lemma}
\label{lemma:3cyclesgenAn}
Let $X$ be the set of 3-cycles in $G=A_n$ for $n\geq 3$. Then $G=\la X\ra$, and if $n\geq 5$ then all 3-cycles are conjugate.
\end{lemma}
\begin{proof}
By sheet 2 Q7, every element of $A_n$ can be written as a product of an even number of transpositions. Hence it suffices to prove that $(a,b)(c,d)$ can be written as a product of 3-cycles.
\begin{enumerate}
\item[$1^\circ$:] $(a,b)=(c,d)$, then $(a,b)(c,d)=1=(1,2,3)^3$.
\item[$2^\circ$:] $|\{a,b\}\cap \{c,d\}|=1$. WLOG $a=c$. Then $(a,b)(c,d)=(a,b)(a,d)=(a,d,b).$
\item[$3^\circ$:] $\{a,b\}\cap \{c,d\}=\varnothing$, then $(a,b)(c,d)=(a,b,c)(b,c,d).$
\end{enumerate}

Now $G$ acts on $X$ by conjugation. It suffices to show $\orb_G((1,2,3))=X$. So let $(a,b,c)\in X$ with $a,b,c$ distinct. We want to find $g\in G:g(1,2,3)g^{-1}=(a,b,c)$.
\begin{enumerate}
\item[$1^\circ$:] $\{1,2,3\}\cap \{a,b,c\}=\varnothing$. Set $g=(1,2)(1,a)(2,b)(3,c)$. We add $(1,2)$ just to make $g$ even, and it doesn't effect since disjoint cycles commute and $(1,2)(a,b,c)(1,2)^{-1}=(a,b,c)(1,2)(1,2)^{-1}=(a,b,c)$.
\item[$2^\circ,\ 3^\circ$:] Similar.
\end{enumerate}
\end{proof}

\begin{lemma}
Let $n\geq 5$ and $\sigma\in A_n$. Then $\exists$ a conjugate $\sigma'\neq\sigma$ and some $i\in\{1,\ldots,n\}$ such that $\sigma(i)=\sigma'(i)$.
\end{lemma}
\begin{proof}
Let $r$ be the length of the longest cycle in $\sigma$. WLOG, we can write $\sigma=(1,2,\ldots,r)\pi$ for some $\pi\in S_n$ with $\pi$ disjoint from $(1,\ldots,r)$ and being a product of cycles of length $\leq r$.
\begin{enumerate}
\item[$1^\circ$:] $r\geq 3$. Then set $g=(3,4,5)$ and $\sigma'=g\sigma g^{-1}=g(1,\ldots,r)g^{-1}g\pi g^{-1}=(1,2,4,\ldots)g\pi g^{-1}$. So $\sigma(1)=\sigma'(1)=2$ but $\sigma(2)=3\neq 4=\sigma'(2)$.
\item[$2^\circ$:] $r\leq 2$. Left as an exercise.
\end{enumerate}
\end{proof}

\begin{remark}
\begin{enumerate}
\item Recall if $N\unlhd G$ and $H\leq G$ then $H\cap N\unlhd H$ (2nd isomorphism theorem).
\item Exercise: if $i\in\{1,\ldots,n\}$ then $\stab_{A_n}(i)\cong A_{n-1}$.
\end{enumerate}
\end{remark}

\begin{thm}
$A_n$ is simple for $n\geq 5$.
\end{thm}
\begin{proof}
Suppose $G=A_n$ is not simple and let $N\unlhd G$. We prove by induction on $n$ with base case $n=5$. By lemma above, for $1\neq \sigma\in N,\ \exists i\in\{1,\ldots,n\}$ and $g\sigma g^{-1}\neq \sigma:$
\[
\left(g\sigma g^{-1}\right)^{-1} (\sigma) (i)=i,
\]
so
\[
1\neq \left(g\sigma g^{-1}\right)^{-1} (\sigma) \in N\cap \stab_G(i) \unlhd \stab_G(i).
\]
So by induction hypothesis which says $\stab_G(i)\cong A_{n-1}$ is simple, $N\cap \stab_G(i)$ can only be the whole group $\stab_G(i)$. So $\stab_G(i)\leq N$, hence $N$ contains a 3-cycle. But then by previous lemmas, $N$ contains all 3-cycles. But $A_n$ is generated by 3-cycles, so $A_n\leq N$. Hence $A_n=N$, a contradiction.
\end{proof}

\begin{flushright}
\textit{Week 7, lecture 1 starts here}
\end{flushright}

\section{Classifying groups of small order}
\subsection{Semidirect product}
\begin{defn}
Let $H,K$ be groups. Define a binary operation $\cdot:(H\times K)\times (H\times K)\rightarrow H\times K$ by $(h_1,k_1)\cdot (h_2,k_2)=(h_1h_2,k_1k_2)$. Then $(H\times K,\cdot)$ is a group, called the \textit{direct product} of $H$ and $K$, denoted usually simply $H\times K$.
\end{defn}
\begin{remark}
\begin{enumerate}
\item One can generalise this definition to product of more than 2 groups.
\item The identity of $G_1\times \cdots \times G_t$ is $\left(1_{G_1},\ldots,1_{G_t}\right)$, and $(g_1,\ldots,g_t)^{-1}=\left(g_1^{-1},\ldots,g_t^{-1}\right)$.
\item $H\times K\cong K\times H$.
\end{enumerate}
\end{remark}

\begin{lemma}
\label{lemma:GisHKthenGcongHKifHKnormalanditsc1}
Let $H,K\unlhd G$ with $H\cap K=\{1_G\}$ and $G=HK$. Then
\begin{enumerate}
\item $hk=kh \ \forall h\in H,k\in K$.
\item $G\cong H\times K$.
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{enumerate}
\item Let $h\in H,k\in K$. Note $hk=kh \Leftrightarrow hkh^{-1}k^{-1}=1$. Since $H\unlhd G,\ kh^{-1}k^{-1}\in H$, so $hkh^{-1}k^{-1}\in H$. By symmetry of $H$ and $K,\ hkh^{-1}k^{-1}\in K$ as well, so $hkh^{-1}k^{-1}=1$ as desired.
\item Define $\varphi:H\times K\rightarrow HK=G$ by $(h,k)\mapsto hk$. Sanity check: if $(h_1,k_1),(h_2,k_2)\in H\times K$ then $\varphi((h_1,k_1)(h_2,k_2))=\varphi((h_1h_2,k_1k_2))=h_1h_2k_1k_2=h_1k_1h_2k_2=\varphi((h_1,k_1))\varphi((h_2,k_2))$. It immediately follows from assumption that $\varphi$ is surjective. Now if $(h,k)\in\ker\varphi$ then $h=k^{-1}\in H\cap K=\{1\}$, so $h=k=1$ and $hk=1$, i.e. $\ker\varphi=\{1\}$ which implies $\varphi$ is injective. 
\end{enumerate}
\end{proof}

\begin{remark}
The hypotheses of this lemma are not too bad to work with. Lagrange's theorem allows us to study $H\cap K$, Proposition \ref{prop:orderofHK} allows to study $HK$, and Sylow theorems say a lot about normality.
\end{remark}

\begin{defn}
An isomorphism $\phi:G\rightarrow G$ is an \textit{automorphism} of $G$. The set $\Aut(G):=\{\phi:\phi\text{ an automorphism}\}$ is a group under composition, called the \textit{automorphism group} of $G$.
\end{defn}
\begin{example}
\label{example:auto}
\begin{enumerate}
\item $\id:G\rightarrow G$ is an automorphism.
\item If $G=C_p$ where $p$ is prime, then $f_e:G\rightarrow G:x_i\mapsto x^{ie}\in \Aut(G)$ for $1\leq e\leq p-1$. Furthermore, this is in fact all the automorphisms and $\Aut(C_p)\cong C_{p-1}$ (see sheet 4 Q8).
\item If $K\unlhd G$ and $g\in G$, then $c_g:K\rightarrow K:x\mapsto gxg^{-1}\in\Aut(K)$.
\end{enumerate}
\end{example}

\begin{defn}
Let $H,K$ be groups and $\phi:H\rightarrow \Aut(K)$ a homomorphism. For $h\in H$, write $\phi_h$ in place of $\phi(h)$. Define a binary operation $\ast:(H\times K)\times(H\times K)\rightarrow H\times K$ by $(h_1,k_1)\ast (h_2,k_2) = \left(h_1h_2,\phi_{h_2^{-1}} (k_1)k_2\right)$. Then $(H\times K,\ast)$ is a group, called the \textit{semidirect product} of $H$ and $K$ with respect to $\phi$, denoted $H\ltimes_\phi K$.
\end{defn}

\begin{remark}[Defence of the definition]
This is not as weird as it looks. If $x,y\in G$ then $xy=yc_{y^{-1}}(x)$ where $c_y$ is as in Example \ref{example:auto}.3 above. Also, this really is a generalisation of the direct product. To see this, define $\phi_h$ to be $\id_K \ \forall h\in H$.
\end{remark}

\begin{flushright}
\textit{Week 7, lecture 2 starts here}
\end{flushright}

\begin{example}
\begin{enumerate}
\item Inversion homomorphism: let $H=\la x\ra$ with $|x|=2$ and $K$ be abelian. Define $\phi:H\rightarrow\Aut(K)$ by $\phi_{1_H}=\id_K$ and $\phi_x(k)=k^{-1}$.

Check $\phi_x$ is an automorphism: indeed $\phi_x\in\Aut(K)$ since it's clearly bijective and as $K$ is abelian, $\phi_x(k_1k_2)=k_2^{-1}k_1^{-1}=k_1^{-1}k_2^{-1}=\phi_x(k_1)\phi_x(k_2)$.

Check $\phi$ is a homomorphism, i.e. $\phi_{h_1h_2}=\phi_{h_1}\circ\phi_{h_2} \ \forall h_1,h_2\in H$, which is not difficult to show.
\item Conjugation homomorphism: let $G$ be finite and $K\unlhd G,H\leq G$. Define $\phi$ by $\phi_h(k)=hkh^{-1}$. Again it's not difficult to do the two sanity checks.
\end{enumerate}
\end{example}

\begin{lemma}[General form of Lemma \ref{lemma:GisHKthenGcongHKifHKnormalanditsc1}]
\label{lemma:generalformofHKcongHsK}
If $H\leq G,K\unlhd G$ with $H\cap K=\{1_G\}$ and $G=HK$, then $G\cong H\ltimes_\phi K$ where $\phi$ is conjugation homomorphism.
\end{lemma}

\begin{proof}
Again it suffices to show $f:H\ltimes_\phi K\rightarrow G:f((h,k))=hk$ is an isomorphism. Let $(h_1,k_1),(h_2,k_2)\in H\ltimes_\phi K$, then
\[
\begin{aligned}
f((h_1,k_1)(h_2,k_2))&=f((h_1h_2,\phi_{h_2^{-1}}(k_1)k_2))=h_1h_2\phi_{h_2^{-1}}(k_1)k_2\\
&=h_1h_2(h_2^{-1}k_1h_2)k_2=h_1k_1h_2k_2=f((h_1,k_1))f((h_2,k_2)).
\end{aligned}
\]
To show $f$ is bijective is similar to proof of Lemma \ref{lemma:GisHKthenGcongHKifHKnormalanditsc1}.
\end{proof}

\begin{example}[Dihedral groups as semidirect products]
Recall Definition \ref{defn:dihedralgp}, write $G=D_{2n}=\la\sigma,\tau\ra$, and let $C_2\cong H:=\la \tau\ra\leq G,\ C_n\cong K:=\la \sigma\ra\unlhd G$. Proposition \ref{prop:orderofHK} says
\[
|HK|=\frac{|H||K|}{|H\cap K|}=\frac{2\times n}{1}=2n=|G|
\]
since if $H\cap K\neq\{1\}$ then it would have to be $H$ since $|H|=2$. Thus $G=HK$, so by previous lemma $G\cong H\ltimes_\phi K$ where $\phi$ is conjugation homomorphism.

Note that
\[
\phi_\tau(\sigma)=\tau\sigma\tau^{-1}=(\tau(1),\tau(2),\ldots,\tau(n))=(n,n-1,\ldots,1)=\sigma^{-1}
\]
and in general $\phi_\tau(\sigma^i)=\sigma^{-i}$, so $\phi$ is also inversion homomorphism.
\end{example}

\begin{lemma}[Generalising example above]
\label{lemma:spottingD2n}
Let $G$ be nonabelian and finite. If \begin{itemize}
\item $G$ has a cyclic subgroup $K$ of order $\frac{|G|}{2}=:n$,
\item $G\backslash K$ has an element $x$ of order 2, and
\item the only $i\in\{1,\ldots,n-1\}:i^2\equiv 1\Mod n$ are $1$ and $n-1$,\hfill ($\dagger$)
\end{itemize}
then $G\cong D_{2n}$.
\end{lemma}

\begin{proof}
First note that $\dagger$ is satisfied when $n=6,\ n=p$ or $n=p^2$ where $p$ is prime.

\begin{flushright}
\textit{Week 7, lecture 3 starts here}
\end{flushright}

Set $H=\la x\ra\leq G$ and note that $K\unlhd G$ since $[G:K]=2,\ H\cap K=\{1_G\}$ since $x\in G\backslash K$ and
\[
|HK|=\frac{|H||K|}{|H\cap K|}=2n=|G|,
\]
so $G=HK$. Recall Lemma \ref{lemma:generalformofHKcongHsK}, assumptions of which are all satisfied. It remains to show that conjugation homomorphism $\phi$ is equal to inversion homomorphism here by example above, i.e. showing $\phi_x(k)=k^{-1} \ \forall k\in K$. Since $K$ is cyclic of order $n$, one can write $K=\la y\ra$ with $|y|=n$. By exercises below, it suffices to show $\phi_x(y)=y^{-1}$. Note that $xyx^{-1}\in \la y\ra$ since $K\unlhd G$, i.e. $xyx^{-1}=y^i$ for some $i\in\{1,\ldots,n-1\}$. Since $\phi_{1_H}=\id_K$, one has
\[
y=\phi_{1_H}(y)=\phi_{x^2}(y)=(\phi_x\circ \phi_x)(y)=\phi_x(\phi_x(y))=\phi_x(y^i)=\phi_x(y)^i=(xyx^{-1})^n,
\]
so $y=y^{i^2}$, i.e. $y^{i^2-1}=1_G$. By Lemma \ref{lemma:orderlem}, $n\mid i^2-1$, i.e. $i^2\equiv 1\Mod n$, so by assumption $i=1$ or $n-1$. One now has that $\phi_x(y)=y$ or $y^{-1}$, but if $xyx^{-1}=y$ then $xkx^{-1}=k \ \forall k\in K$, i.e. $\phi$ is trivial homomorphism, which implies $G=H\ltimes_\phi K\cong H\times K\cong C_2\times C_n$ is abelian, contradicting assumption. So $\phi_x(y)=y^{-1}$, inversion homomorphism.
\end{proof}

\begin{exe}
\begin{enumerate}
\item If $H=\la A\ra,\ K=\la B\ra$, show $hkh^{-1}=k \ \forall h\in H,k\in K\Leftrightarrow aba^{-1}=b \ \forall a\in A,b\in B$.
\item If $H=\la x\ra$ with $|x|=2$ and $K=\la B\ra$ is abelian, show $xkx^{-1}=k^{-1} \ \forall k\in K\Leftrightarrow xbx^{-1}=b^{-1} \ \forall b\in B$.
\end{enumerate}
\end{exe}

\subsection{Semidirect product of an abelian group and a cyclic group}
In this section, we fix a finite group $G$ with an abelian subgroup $K$ of odd order $\frac{|G|}{2}$ (so we know it's normal) and let $H=\la x\ra\in\Syl_2(G)$ with $|x|=2$ (as $|G|=2\times$odd number).

\begin{notation}
For $v\in K,$ write $[v,x]:=vxv^{-1}x^{-1}$ (the commutator).
\end{notation}

\begin{lemma}[Fitting's]
Write $[K,x]:=\la[v,x]:v\in K\ra$. One has
\begin{enumerate}
\item $xkx^{-1}=k^{-1} \ \forall k\in [K,x]$,
\item $K\cong [K,x]\times C_K(x)$,
\item $G\cong (H\ltimes_\phi [K,x])\times C_K(x)$ where $\phi$ is inversion homomorphism.
\end{enumerate}
\end{lemma}
\begin{proof}
\begin{enumerate}
\item It suffices to show it for $k=[v,x]$, a generator of $[K,x]$. Since $|x|=2$, one has
\[
x[v,x]x^{-1}=xvxv^{-1}x^{-1}x^{-1}=xvx^{-1}v^{-1}=[v,x]^{-1}.
\]

\begin{flushright}
\textit{Week 8, lecture 1 starts here}
\end{flushright}

\item First note that for $v,w\in K,$
\[
\begin{aligned}
[vw,x]&=(vw)x(vw)^{-1}x^{-1}=vwxw^{-1}v^{-1}x^{-1}\\
&=vw\left(xw^{-1}x^{-1}\right)\left(xv^{-1}x^{-1}\right) \\
&=[v,x][w,x],
\end{aligned}
\]
and $[v,x]=1$ iff $v$ and $x$ commute, i.e. $v\in C_K(x)$.

Now define $f:K\rightarrow [K,x]$ by $f(k)=[k,x]$. $f$ is a homomorphism with $\ker f=C_K(x)$ by above. We claim \begin{enumerate}
\item $C_K(x),[K,x]\unlhd K$. This is trivial since $K$ is abelian.
\item $C_K(x)\cap [K,x]=\{1\}$. Indeed, if $a\in C_K(x)\cap [K,x]$, then $a=xax^{-1}=a^{-1}$ by part 1, so $|a|=1$ or 2. But since $|K|$ is odd, by Lagrange's $|a|$ must be 1, so $a=1$.
\item $K=[K,x]C_K(x)$. Indeed, by 1st isomorphism theorem $|K|=|\im f||\ker f|=|[K,x]||C_K(x)|$, hence by \ref{prop:orderofHK} one has
\[
|[K,x]C_K(x)|=\frac{|[K,x]||C_K(x)|}{|[K,x]\cap C_K(x)|}=|[K,x]||C_K(x)|=|K|.
\]
\end{enumerate}
So by Lemma \ref{lemma:GisHKthenGcongHKifHKnormalanditsc1} one has the desired.
\item Left as an exercise, see sheet 4 Q14.
\end{enumerate}
\end{proof}

\subsection{Infinite families}
\subsubsection{Abelian groups}
\begin{thm}[Fundamental theorem of finite abelian groups]
\label{thm:FTFAG}
Let $G$ be a finite abelian group of order $n$. Then $\exists$ divisors $d_1\mid \cdots \mid d_t$ of $n$ such that $G\cong C_{d_1}\times\cdots\times C_{d_t}$.
\end{thm}
\begin{proof}
See MA251.
\end{proof}

\begin{example}
The abelian groups of order 8 are $C_8,\ C_2\times C_4,\ C_2\times C_2\times C_2$.
\end{example}

\subsubsection{Groups of order $p,p^2,2p$ where $p$ prime}
\begin{lemma}
$|G|=p\Rightarrow G\cong C_p$.
\end{lemma}
\begin{proof}
Note that by Lagrange's, any $x\in G\backslash \{1\}$ has $|x|=p$, so $G=\la x\ra\cong C_p$.
\end{proof}

\begin{lemma}
$|G|=p^2\Rightarrow G\cong C_{p^2}$ or $C_p\times C_p$.
\end{lemma}
\begin{proof}
This follows immediately from Corollary \ref{coro:orderp2gpabelian} and Theorem \ref{thm:FTFAG}.
\end{proof}

\begin{lemma}
If $p$ is odd, then $|G|=2p\Rightarrow G\cong C_{2p}$ or $D_{2p}$.
\end{lemma}
\begin{proof}
If $G$ is abelian then $G\cong C_{2p}$ by \ref{thm:FTFAG}. If $G$ is nonabelian and let $K\in\Syl_p(G),\ H=\la x\ra \in\Syl_2(G)$ where $|x|=2$. Then
\begin{enumerate}
\item $|K|=\frac{|G|}{2}=p$, so $K\cong C_p$,
\item $x\in G\backslash K$,
\item If $i^2\equiv 1\Mod p$ then $i\equiv \pm 1\Mod p$ since $\Z/p\Z$ is a field,
\end{enumerate}
so by Lemma \ref{lemma:spottingD2n} one has $G\cong D_{2p}$.
\end{proof}

\subsubsection{Groups of order $2p^2$ where $p$ odd prime}
\begin{defn}
Let $K=C_p\times C_p,\ H=C_2$ and $\phi:H\rightarrow\Aut(K)$ be inversion homomorphism. $H\ltimes_\phi K$ is called the \textit{generalised dihedral group} of order $2p^2$ and denoted $GD_{2p^2}$.
\end{defn}

\begin{lemma}
$|G|=2p^2\Rightarrow G\cong$ either $C_{2p^2},\ C_p\times C_{2p},\ GD_{2p^2},\ D_{2p^2}$ or $C_p\times D_{2p}$.
\end{lemma}
\begin{proof}
\begin{enumerate}
\item If $G$ is abelian then $G\cong C_{2p^2}$ or $C_p\times C_{2p}$ by \ref{thm:FTFAG}.

\item If $G$ is nonabelian, let $K\in\Syl_p(G)$ and $H=\la x\ra\in\Syl_2(G)$ where $|x|=2$. Then $K\cong$ either $C_{p^2}$ or $C_p\times C_p$.
\begin{enumerate}
\item If $K\cong C_{p^2}$ then similarly the three conditions of \ref{lemma:spottingD2n} are satisfied and $G\cong D_{2p^2}$.
\begin{flushright}
\textit{Week 8, lecture 2 starts here}
\end{flushright}
\item If $K\cong C_p\times C_p$ then by Fitting's lemma, $G\cong (H\ltimes_\phi [K,x])\times C_K(x)$ where $\phi$ is inversion homomorphism. By Lagrange's, $|C_K(x)|$ is either $1,p$ or $p^2$. But if $|C_K(x)|=p^2$ then $K=C_K(x)$, i.e. $kx=xk \ \forall k\in K$, contradicting $G$ being abelian since it is generated by $K$ and $x$. 
\begin{enumerate}
\item If $|C_K(x)|=1$ then $C_K(x)=\{1\}$. By Fitting's lemma, $K\cong [K,x]\times C_K(x)\cong [K,x]$ and so $G\cong H\ltimes_\phi K\cong C_2\ltimes_\phi(C_p\times C_p)\cong GD_{2p^2}$.
\item If $|C_K(x)|=p$ then $C_K(x)\cong C_p$ and so $[K,x]\cong C_p$, therefore $G\cong (H\ltimes_\phi [K,x]) \times C_K(x)\cong (C_2\ltimes_\phi C_p) \times C_p\cong D_{2p}\times C_p$.
\end{enumerate}
\end{enumerate}
\end{enumerate}
\end{proof}

\subsubsection{Groups of order $pq$ where $p,q$ prime with $p<q$ and $p\nmid (q-1)$}
\begin{lemma}
Let $p,q$ be as above. Then $|G|=pq\Rightarrow G\cong C_{pq}$.
\end{lemma}
\begin{proof}
By Sylow theorems, $|\Syl_q(G)|\equiv 1\Mod q$ and $|\Syl_q(G)|\mid \frac{|G|}{|G|_q}=p$. Since $q>p$ and $p\not\equiv 1\Mod q,\ |\Syl_q(G)|=1$. Similarly, $|\Syl_p(G)|=1$. Write $\Syl_p(G)=\{P\}$ and $\Syl_q(G)=\{Q\}$. Then $P,Q\unlhd G,\ P\cap Q=\{1\}$ and $G=PQ$ since $|PQ|=\frac{|P||Q|}{|P\cap Q|}=pq=|G|$. So by Lemma \ref{lemma:GisHKthenGcongHKifHKnormalanditsc1} and Theorem \ref{thm:FTFAG}, $G\cong P\times Q\cong C_p\times C_q\cong C_{pq}$.
\end{proof}

\subsection{2 missing pieces}
\subsubsection{Groups of order 8}
\begin{defn}
Let $i,j,k$ be indeterminates and define
\[
Q_8:=\{\pm1,\pm i,\pm j,\pm k\}\subseteq \mathbb R[i,j,k].
\]
Define binary operation $\cdot:Q_8\times Q_8\rightarrow Q_8$ by
\begin{enumerate}
\item $1\cdot g=g\cdot 1:=g$ and $(-1)\cdot g=g\cdot(-1):=-g \ \forall g\in Q_8$.
\item $i\cdot j:=k,\ j\cdot k:=i,\ k\cdot i:=j$.
\item $j\cdot i:=-k,\ k\cdot j:=-i,\ i\cdot k:=-j$.
\item $(\pm 1)^2=1,\ g^2:=-1 \ \forall g\in Q_8\backslash\{\pm1\}$.
\end{enumerate}
$(Q_8,\cdot)$ is then a group with its full Cayley table determined, called the \textit{quaternion group}.
\end{defn}

\begin{remark}
\begin{enumerate}
\item $Z(Q_8)=\{\pm1\}$.
\item $Q_8$ has 1 element of order 2 ($-1$) and 6 elements of order 4 ($\pm i,\pm j,\pm k$).
\item $Q_8=\la i,j\ra=\la i,k\ra=\la j,k\ra$.
\end{enumerate}
\end{remark}

\begin{lemma}
$|G|=8\Rightarrow G\cong$ either $C_8,\ C_2\times C_4,\ C_2\times C_2\times C_2,\ D_8$ or $Q_8$.
\end{lemma}
\begin{proof}
\begin{enumerate}
\item If $G$ is abelian then by \ref{thm:FTFAG} $G\cong$ either $C_8,\ C_2\times C_4$ or $C_2\times C_2\times C_2$.
\item If $G$ is nonabelian, then of the 7 elements of order $>1$, none has order 8 (since then $G$ would be $C_8$) and at least one has order $\neq 2$ (since if all elements have order 2, $G$ would be abelian), so by Lagrange's there must $\exists u\in G:|u|=4$. Let $K=\la u\ra$ and $v\in G\backslash K$ with minimal order. One then has $G=\la u,v\ra$. We claim $vuv^{-1}=u^{-1}$. Indeed, $vuv^{-1}\in K=\{1,u,u^2,u^{-1}\}$ since $K\unlhd G$. We know $|vuv^{-1}|=|u|=4$ and $|u^2|=2$, so $vuv^{-1}$ is either $u$ or $u^{-1}$. But if $vuv^{-1}=u$ then $G$ would be abelian, a contradiction. Now
\begin{enumerate}
\item If $|v|=2$ then conditions of Lemma \ref{lemma:spottingD2n} are satisfied, so $G\cong D_8$.
\item If $|v|=4$, note that $G=K\sqcup vK$ and all elements of $vK$ have order 4, so $G$ has 1 element of order 2 ($u^2$) and 6 elements of order 4. It follows that $g^2=u^2 \ \forall g\in G:|g|=4$, since $g^2$ has order 2 and $u^2$ is the only such element. Now if we see $G$ as $\{1,u^2,u^{\pm 1},v^{\pm 1},(uv)^{\pm 1}\}$ we have $G\cong Q_8$.
\end{enumerate}
\end{enumerate}
\end{proof}

\begin{flushright}
\textit{Week 8, lecture 3 starts here}
\end{flushright}

\subsubsection{Groups of order 12}
\begin{defn}
\label{defn:dic}
Let $H=C_4=\la x\ra$ where $|x|=4$ and $K=C_3$. Define $\phi:H\rightarrow\Aut(K)$ by $\phi_{x^i}(k)=k^{(-1)^i}$. The group $H\ltimes_\phi K$ is called the \textit{dicyclic group} or order 12, denoted $\Dic_{12}$.
\end{defn}

\begin{lemma}
$|G|=12\Rightarrow G\cong$ either $C_{12},\ C_2\times C_6,\ \Dic_{12},\ A_4$ or $D_{12}$.
\end{lemma}
\begin{proof}
\begin{enumerate}
\item If $G$ is abelian then by \ref{thm:FTFAG} $G\cong$ either $C_{12}$ or $C_2\times C_6$.
\item If $G$ is nonabelian, then
\begin{enumerate}
\item If $G$ has an element $a$ of order 6, let $K=\la a\ra$. Then all subgroups of $K$ are normal (see sheet 3 Q9). Now
\begin{enumerate}
\item If $G\backslash K$ has an element of order 2 then conditions of \ref{lemma:spottingD2n} hold, so $G\cong D_{12}$.
\item If $G\backslash K$ has no element of order 2, then let $P\in\Syl_2(G)$. We know $P\not\leq K$by Lagrange's, so choose $x\in P\backslash K$. $|x|$ can only be 4 since it cannot be 1 or 2. Let $H=\la x\ra$ and $K_1=\la a^2\ra$. So $|H|=4,\ |K_1|=3$, and so conditions of \ref{lemma:generalformofHKcongHsK} hold and $G\cong H\ltimes_\phi K_1$ where $\phi$ is conjugation homomorphism. We claim that in this case, $\phi$ is the same as the $\phi$ defined in \label{defn:dic} and so that $G\cong\Dic_{12}$. Indeed, let $k\in\{a^2,a^{-2}\}\subseteq K_1$. Then $G=\la x,k\ra$ and $xkx^{-1}$ is either $k$ or $k^{-1}$ since $K_1\unlhd G$. But $xkx^{-1}\neq k$ since $G$ is nonabelian. So $\phi_x(k)=xkx^{-1}=k^{-1}$ and hence $\phi_{x^i}(k)=k^{(-1)^i}$ since $\phi$ is a homomorphism.
\end{enumerate}
\item If $G$ has no element of order 6, then let $P=\la x\ra\in \Syl_3(G)$ where $|x|=3$. By Sylow theorems, $|\Syl_3(G)|\in\{1,4\}$. By \ref{tool:elementorder}.3, $|C_G(x)|$ is odd, and since $x\in C_G(x),\ |C_G(x)=3$ and $|F_3(G)|\geq |G:C_G(x)|=4$. By \ref{coro:consequence2Sylow}.3, $|F_3(G)|=2|\Syl_3(G)|$, so $|\Syl_3(G)|\geq 2$, hence $|\Syl_3(G)|=4$ and $P\ntrianglelefteq G$. Now let $G$ act on $X=G/P$ by left multiplication. By sheet 2 Q9, $\ker(G,X,\cdot)\leq P$. By Lagrange's, $\ker(G,X,\cdot)$ is either trivial or $P$, but kernels are normal, so $\ker(G,X,\cdot)$ is trivial and the action is faithful, so by \ref{prop:faithfulisosgpsymX}, $G\cong$ a subgroup of $S_4$. The only subgroup of $S_4$ of order 12 is $A_4$.
\end{enumerate}
\end{enumerate}
\end{proof}

\subsection{Final theorem of the chapter}

\begin{thm}
The only simple group of order 60 is $A_5$.
\end{thm}
\begin{proof}
Let $G$ be a simple group of order $60=2^2\times3\times5$. We claim:
\begin{enumerate}
\item If $H\lneq G$ then $|H|\leq 12$.

Indeed, since $G$ is simple, $G$ acts faithfully on $X=G/H$ (as $\ker(G,X,\cdot)\leq H\lneq G$), so by \ref{prop:faithfulisosgpsymX}, $G\cong$ a subgroup of $S_{|G:H|}$ and hence $|G|\mid |G:H|!$. Since $|G|=60,\ |G:H|\geq 5$, so $|H|\leq 12$, i.e. $G$ has no subgroup of index 4 or less.
\item If $P_1,P_2\in\Syl_2(G)$ and $P_1\cap P_2\neq \{1\}$, then $H=\la P_1\cup P_2\ra\Rightarrow |H|=12$.

Indeed, let $x\in P_1\cap P_2\backslash\{1\}$. Then $x\in Z(P_1)\cap Z(P_2)$ since $|P|=4$ and all groups of order 4 are abelian. So $H=\la P_1\cup P_2\ra \leq C_G(x)$. Since $G$ is simple, $Z(G)=\{1\}$, so $C_G(x)\lneq G$ and so $H\lneq G$. By claim 1 $|H|\leq 12$.

\begin{flushright}
\textit{Week 9, lecture 1 starts here}
\end{flushright}

Now, $4\mid |H|$ since $|P_1|=4$. Also, $|\Syl_2(H)|\equiv 1\Mod 2$ by Sylow theorems, i.e. it's odd, and by \ref{coro:consequence1Sylow} $|\Syl_2(H)|\mid |H|$, so $4\times(\text{an odd number})\mid |H|$. But as $P_1,P_2\leq H,\ |\Syl_2(H)|\geq 2$, so $|H|\geq 12$. We conclude that $|H|=12$.
\item If $P_1\cap P_2=\{1\} \ \forall P_1\neq P_2,\ P_1,P_2\in\Syl_2(G)$ then $H=N_G(P)$ has order 12 $\forall P\in\Syl_2(G)$.

Indeed, again $|\Syl_2(G)|=|G:N_G(P)|$ is odd, so $|\Syl_2(G)|\in\{1,3,5,15\}$. It's not 1 since $G$ is simple, it's not 3 by claim 1, so $|\Syl_2(G)|\in\{5,15\}$. Suppose it's 15. Note that $|\Syl_5(G)|=6$, and by \ref{coro:consequence2Sylow}, since $|G|_5=5,\ |F_5(G)|=6\times(4-1)=24$. By the assumption that $P_i\backslash\{1\}$ are disjoint,
\[
|F_2(G)|=\left|\bigcup_{P\in\Syl_2(G)} P\backslash \{1\}\right|=15\times(4-1)=45,
\]
so we have at least 69 elements in a group of order 60, an absurdity. Hence $|\Syl_2(G)|=5$ and $|N_G(P)|=12$. Combined with claim 2, this means $G$ always has a subgroup of index 5.
\end{enumerate}
Now let $H\leq G$ be a subgroup of index 5, then again by argument in proof of claim 1, $G\cong$ a subgroup of $S_5$, but the only subgroup of $S_n$ of order $\frac{n!}{2}$ is $A_n$.
\end{proof}

\section{Soluble group and Jordan–Hölder theorem}
\subsection{Composition series}
\begin{notation}
$H\lneq G$ means $H$ is a proper subgroup and $H\pns G$ means $H$ is a proper normal subgroup.
\end{notation}
\begin{defn}
A \textit{composition series} for a group $G$ is a series
\[
\{1_G\}=G_0\pns G_1\pns\cdots\pns G_r=G
\]
where the (finite) $r$ is called \textit{length} of the series, such that $G_i/G_{i-1}$ is simple $\forall 1\leq i\leq r$.
\end{defn}
\begin{example}
\label{example:compositionseries}
\begin{enumerate}
\item If $G=D_{2p}$ with the usual generators $\sigma,\tau$, then
\[
G_0=\{1\},\ G_1=\la\sigma\ra,\ G_2=G
\]
is a composition series since $C_2$ and $C_p$ are simple.
\item If $G=S_n$ where $n\geq 5$ then
\[
G_0=\{1\},\ G_1=A_n,\ G_2=G\]
is a composition series.
\item If $G=D_8$ with $\sigma=(1,2,3,4)$ and $\tau=(1,4)(2,3)$, then
\[
G_0=\{1\},\ G_1=\la\sigma^2\ra,\ G_2=\la\sigma^2,\tau\ra,\ G_3=G
\]
is a composition series. We could have also set $G_2=\la\sigma\ra$.
\end{enumerate}
\end{example}

\begin{thm}
\label{thm:everyfingphasacompnseries}
Every finite group has a composition series.
\end{thm}
\begin{proof}
By convention, the trivial group has the composition series $G_0=G$ of length 0. We then proceed to prove by induction on $|G|$ and assume all groups of order $<|G|$ have a composition series.

\begin{flushright}
\textit{Week 9, lecture 2 starts here}
\end{flushright}

If $G$ is simple then $G_0=\{1\},\ G_1=G$ is a composition series, so suppose $G$ is not simple. Then $\exists N: \{1\}\neq N\neq G$ and $N\unlhd G$. By inductive hypothesis, $N$ and $G/N$ both have a composition series, and one writes
\[
\begin{aligned}
\{1_G\}&=N_0\pns N_1\pns \cdots\pns N_r=N \\
\{1_{G/N}\}&=\overline{G_0}\pns\overline{G_1}\pns\cdots\pns\overline{G_s}=G/N
\end{aligned}
\]
Now, by Theorem \ref{thm:correspondence}, $\exists X_i:N\leq X_i\leq G$ and $X_i/N=\overline{G_i}$ for all $i=1,\ldots,s$. Also by Theorem \ref{thm:3rdiso}, one has
\[
\frac{X_i/N}{X_{i-1}/N}\cong \frac{X_i}{X_{i-1}},
\]
which is simple. Now define
\[
G_i:=\left\{ \begin{aligned}
& N_i,\quad && 1\leq i\leq r \\
& X_{i-r},\quad && r+1\leq i \leq r+s
\end{aligned} \right.
\]
and note that $X_r=N$ since $\{1_{G/N}\}=N/N$, so
\[
G_0\pns G_1\pns\cdots\pns G_{r+s}=G
\]
is a composition series.
\end{proof}

\begin{coro}[Direct byproduct of proof but useful to write down]
\label{coro:adjoinseriesofnormalandquotient}
Let $G$ be a finite group, $N\unlhd G$, and
\[
\begin{aligned}
\{1_G\}&=N_0\pns N_1\pns\cdots\pns N_r=N \\
\{1_{G/N}\}&=N/N\pns X_1/N\pns\cdots\pns X_s/N=G/N
\end{aligned}
\]
be composition series for $N$ and $G/N$ where $N\leq X_i\leq G$, then
\[
G_i:=\left\{ \begin{aligned}
&N_i,\quad &&1\leq i\leq r \\
&X_{i-r},\quad &&r+1\leq i \leq r+s
\end{aligned} \right.
\]
yields a composition series.
\end{coro}

\begin{example}
Recall \ref{example:compositionseries}.3 in which we have two different composition series for $D_8$. But they are not that different after all: length is both 3 and all $G_i/G_{i-1}\cong C_2$ in both cases. Let's codify this.
\end{example}

\begin{defn}
Let
\[
\tag{I}
\{1_G\}=A_0\pns\cdots\pns A_r=G
\]
\[
\tag{II}
\{1_G\}=B_0\pns\cdots\pns B_s=G
\]
be 2 composition series for a group $G$. We say (I) and (II) are \textit{equivalent} (and write (I) $\sim$ (II)) if $r=s$ and $\exists$ a bijection
\[
f:\{A_i/A_{i-1} : 1\leq i\leq r\} \rightarrow \{B_j/B_{j-1}:1\leq j\leq s\}
\]
such that $A_i/A_{i-1}\cong f(A_i/A_{i-1})$.
\end{defn}

\begin{thm}[Jordan–Hölder]
Any two composition series of a finite group are equivalent.
\end{thm}
\begin{proof}
Let two composition series (I) and (II) of a group $G$ be as above, WLOG assume $r\leq s$ and do induction on $r$. Base case $r=0$ is trivial so suppose $r>0$ and statement is true for smaller $r$.
\begin{enumerate}
\item[$1^\circ$] $A_{r-1}=B_{s-1}$ are the same group, then the two series are equivalent by inductive hypothesis.
\item[$2^\circ$] $A_{r-1}\neq B_{s-1}$. The idea is to construct two new composition series to `link' the current ones together. Denote $A_{r-1}$ by $A$ and $B_{s-1}$ by $B$, and let $D:=A\cap B$.

We claim $A\not\leq B$ and $B\not\leq A$. Indeed, suppose $A\leq B$, then $B/A\unlhd G/A$. But $G/A$ is simple, so $B$ is either $A$ or $G$, and since we assume $A\neq B$, it must be $B=G$, but by definition $B\pns G$, a contradiction.

We now claim $D\unlhd A$ and $A/D\cong G/B$ (and symmetrically $D\unlhd B$ and $B/D\cong G/A)$. Indeed, note that by Theorem \ref{thm:2ndiso}.2, since $A\unlhd A_r=G$ so $A\cap B\unlhd B$ and since $B\unlhd B_s=G$ so $A\cap B=D\unlhd A$. Again by \ref{thm:2ndiso}.3,
\[
\frac{A}{D} = \frac{A}{A\cap B} \cong \frac{AB}{B},
\]

\begin{flushright}
\textit{Week 9, lecture 3 starts here}
\end{flushright}

so it remains to prove $AB=G$. Now, $A,B\unlhd G\Rightarrow AB\unlhd G\Rightarrow\frac{AB}{B}\unlhd \frac{G}{B}$, where $G/B$ is simple, so $AB$ is either $B$ or $G$, but $AB\neq B$ since $A\not\leq B$. By Theorem \ref{thm:everyfingphasacompnseries}, $D$ has a composition series
\[
\{1\}=D_0\pns\cdots\pns D_t=D,
\]
and since $A/D$ and $B/D$ are simple, we have two new composition series for $G$
\[
\tag{III}
\{1\}=D_0\pns \cdots\pns D_t\pns A\pns G,
\]
\[
\tag{IV}
\{1\}=D_0\pns \cdots\pns D_t\pns B\pns G,
\]
where (III) $\sim$ (I) and (IV) $\sim$ (II) by case 1, so $r=s=t+2$. Finally, since $G/B\cong A/D$ and $G/A\cong B/D$, we see that (III) $\sim$ (IV) by definition, so (I) $\sim$ (II) by transitivity.
\end{enumerate}
\end{proof}

\begin{defn}
Let $G$ be a finite group and $\{1_G\}=G_0\pns\cdots\pns G_r=G$ a composition series. The factors $G_i/G_{i-1}$  are called the \textit{composition factors} of $G$ and $r$ is called the \textit{composition length} of $G$.

Jordan–Hölder theorem justifies the `the' before the noun defined.
\end{defn}
\begin{remark}
Composition factors don't determine a group, e.g. $D_8$ and $Q_8$.
\end{remark}
\begin{example}
\begin{enumerate}
\item $D_{2p}$ has factors $C_p,C_2$.
\item $S_n$ with $n\geq 5$ has factors $A_n,C_2$.
\item $D_8$ has factors $C_2,C_2,C_2$.
\end{enumerate}
Note that factors of 1 and 3 are all cyclic groups of prime order, this is something special we want to define.
\end{example}
\begin{defn}
A finite group is \textit{soluble} if all composition factors are cyclic of prime order.
\end{defn}

\begin{lemma}
\label{lemma:GsoliffNGNsol}
Let $G$ be a finite group and $N\unlhd G$. Then $G$ is soluble iff $N$ and $G/N$ are soluble.
\end{lemma}
\begin{proof}
See \ref{coro:adjoinseriesofnormalandquotient}.
\end{proof}

\begin{example}
\begin{enumerate}
\item By above, $D_{2n}$ is soluble since $C_n$ and $D_{2n}/C_n=C_2$ are soluble (all abelian groups are soluble).
\item To show $S_4$ is soluble, note that $A_4$ is soluble (proof left as an exercise) and $S_4/A_4=C_2$ is soluble.
\end{enumerate}
\end{example}

\begin{flushright}
\textit{Week 10, lecture 1 starts here}
\end{flushright}

\subsection{Commutator}
\begin{defn}
Let $G$ be a group and $g,h\in G$. The \textit{commutator} of $g,h$ is $[g,h]=ghg^{-1}h^{-1}.$
\end{defn}
\begin{example}
\begin{enumerate}
\item $[g,h]=1_G \Leftrightarrow g,h$ commute, so $[g,h]=1_G \ \forall g,h\in G\Leftrightarrow G$ is abelian.
\item In $A_5$ one has
\[
[(1,2,4)(1,3,5)]=(1,2,4)(1,3,5)(1,4,2)(1,5,3)=(1,2,3),
\]
and in general for $A_n$ with $n\geq 5$ and $a,b,c,d,e\in\{1,\ldots,n\}$ distinct, one has
\[
[(a,b,d)(a,c,e)]=(a,b,c),
\]
so all 3-cycles in $A_n$ are commutators, which therefore generate $A_n$ by Lemma \ref{lemma:3cyclesgenAn}.
\end{enumerate}
\end{example}

\begin{defn}
The subgroup
\[
[G,G]:=\la[g,h]:g,h\in G\ra
\]
is the \textit{commutator subgroup} of $G$.
\end{defn}

\begin{remark}
\begin{enumerate}
\item More generally, we can think about the commutator subgroup of 2 subgroups $H,K\leq G$ with the definition
\[
[H:K]=\la [h,k]:h\in H,k\in K\ra.
\]
\item The commutator subgroup is not necessarily the set of commutators. It's a fair mistake to confuse the two since one has to go to order 96 to find a counterexample of commutator subgroup containing non-commutators.
\end{enumerate}
\end{remark}

\begin{example}
\begin{enumerate}
\item If $G$ is abelian then $[G,G]=\{1_G\}$.
\item $[A_n,A_n]=A_n$ for $n\geq 5$.
\end{enumerate}
\end{example}

\begin{thm}
\label{thm:propofcommsubgp}
\begin{enumerate}
\item $[G,G]\unlhd G$.
\item $G/[G,G]$ is abelian.
\item If $N\unlhd G$ and $G/N$ is abelian then $[G,G]\leq N$, i.e. $[G,G]$ is the smallest normal subgroup $H$ of $G$ with $G/H$ abelian.
\end{enumerate}
\end{thm}
\begin{proof}
\begin{enumerate}
\item It suffices to check that conjugates of one of generators of $[G,G]$, i.e. a commutator, is also a commutator. Let $g,h,k\in G$, then
\[
\begin{aligned}
g[h,k]g^{-1}&=ghkh^{-1}k^{-1}g^{-1}=gh(g^{-1}g)k(g^{-1}g)h^{-1}(g^{-1}g)k^{-1}g^{-1}\\
&=(ghg^{-1})(gkg^{-1})(gh^{-1}g^{-1})(gk^{-1}g^{-1})\\
&=[ghg^{-1},gkg^{-1}].
\end{aligned}
\]
\item Denote $N=[G,G]$ and again let $g,h\in G$. Note that
\[
[g,h]\in N\Rightarrow N[g,h]=N\Rightarrow Nghg^{-1}h^{-1}=N\Rightarrow Ngh=Nhg\Rightarrow (Ng)(Nh)=(Nh)(Ng).
\]
\item By the same (but reverse) argument one has $[g,h]\in N$, so $[G,G]\leq N$.
\end{enumerate}
\end{proof}

\begin{example}
In $G=D_{2n}$ one has $\tau\sigma\tau^{-1}=\sigma^{-1}$, so $[\tau,\sigma]=\sigma^{-2}$. Also $N:=\la\sigma\ra\unlhd G$ and so $K:=\la\sigma^{-2}\ra=\la\sigma^2\ra\leq N$, which implies $K\unlhd G$ (see sheet 3 Q9) and by definition $K\leq [G,G]$.
\begin{enumerate}
\item[$1^\circ$] If $n$ is odd then $K=N$ since $|\sigma^2|=|\sigma|$ (2 and $n$ are coprime), and $|G:K|=2$ with $G/K=C_2$, abelian, so $[G,G]\leq K$, hence $[G,G]=K$.
\item[$2^\circ$] If $n$ is even then $|K|=\frac{n}{2}$ and so $|G/K|=4$ with $G/K$ abelian by \ref{coro:orderp2gpabelian}, so still $[G,G]\leq K$ and hence $[G,G]=K$.
\end{enumerate}
\end{example}

\begin{exe}
\begin{enumerate}
\item Let $\mathbb F$ be a finite field of size at least 4. Show $[G,G]=G$ where $G=SL_2(\mathbb F)$.
\item Show that every element of $A_5$ is a commutator. In fact, if $G$ is a nonabelian simple group and $x\in G$, then $\exists g,h\in G:[g,h]=x$ (this, known as Ore's conjecture, was not proven until 2008).
\end{enumerate}
\end{exe}

\begin{defn}
\label{defn:derivedseries}
Define $G^{(0)}=G,\ G^{(1)}=[G,G]$ and inductively $G^{(i)}=[G^{(i-1)},G^{(i-1)}]$ for $i\geq 2$. The descending series $G^{(0)}\geq G^{(1)}\geq\cdots$ is called the \textit{derived series} of $G$.
\end{defn}
\begin{remark}
\begin{enumerate}
\item $H\leq G\Rightarrow H^{(n)}\leq G^{(n)}$.
\item For $n,m\in\N,\ \left(G^{(n)}\right)^{(m)}=G^{(n+m)}$.
\end{enumerate}
\end{remark}

\begin{flushright}
\textit{Week 10, lecture 2 starts here}
\end{flushright}

\begin{example}
\begin{enumerate}
\item $A_n^{(i)}=A_n \ \forall i\in\N$ for $n\geq 5$.
\item $D_{2n}^{(1)}=\la\sigma^2\ra$ and $D_{2n}^{(2)}=[\la\sigma^2\ra,\la\sigma^2\ra]=\{1_G\}$ since cyclic groups are abelian.
\end{enumerate}
Observe the two patterns: one does not terminate at the trivial group and the other does. Also note that $A_n$ is not soluble for $n\geq 5$ while $D_{2n}$ is. This brings us to the following theorem.
\end{example}

\begin{thm}
\label{thm:GsoliffGneq1}
A finite group $G$ is soluble iff $G^{(n)}=\{1_G\}$ for some $n\in\N$.
\end{thm}
\begin{proof}
\begin{itemize}
\item[$\Rightarrow:$] We prove by induction on $|G|$. The base case is trivial so suppose statement holds for all groups of order $<|G|$ and $|G|>1$. By definition, a composition series $\{1_G\}=G_0\pns\cdots\pns G_r=G$ satisfies $G_i/G_{i-1}$ is cyclic of prime order $\forall 1\leq i\leq r$. In particular, $G_{r-1}\pns G$ and $G/G_{r-1}$ is abelian, so by Theorem \ref{thm:propofcommsubgp} $G^{(1)}\leq G_{r-1}$ and also $G^{(1)}\pns G$. Hence by Lemma \ref{lemma:GsoliffNGNsol}, $G^{(1)}$ is soluble, so by inductive hypothesis $\left(G^{(1)}\right)^n=G^{(1+n)}=\{1_G\}$ for some $n\in\N$.
\item[$\Leftarrow:$] We prove by induction on $n$. The base case $n=1$ implies $G$ is abelian, so soluble. Now suppose statement holds for smaller values of $n$ and denote $[G,G]=G^{(1)}$ by $N$. Then
\[
N^{(n-1)}=\left(G^{(1)}\right)^{(n-1)}=G^{(n)}=\{1_G\},
\]
so $N$ is soluble by inductive hypothesis. Now $G/N$ is abelian by Theorem \ref{thm:propofcommsubgp}, so soluble. Hence $G$ is soluble by Lemma \ref{lemma:GsoliffNGNsol}.
\end{itemize}
\end{proof}

\begin{remark}
\begin{enumerate}
\item In this course, we only defined `soluble' for finite groups. Using theorem above as another characterisation, one can extend the definition to infinite groups.
\item We now have two tools, Lemma \ref{lemma:GsoliffNGNsol} and theorem above, to decide if a given group is soluble.
\end{enumerate}
\end{remark}

\subsection{Examples of a soluble group}
\begin{thm}
\label{thm:subgpofsolissol}
Every subgroup of a finite soluble group is soluble.
\end{thm}
\begin{proof}
This follows immediately from remark after Definition \ref{defn:derivedseries} and Theorem \ref{thm:GsoliffGneq1}.
\end{proof}

\begin{thm}
Every group of order $p^n$ where $p$ prime and $n\in\N$ is soluble.
\end{thm}
\begin{proof}
Let $G$ be such group and we prove by induction on $n$. The base case is trivial since any group of prime order is abelian, so soluble, so suppose $n>1$ and statement is true for any smaller $n$. Denote $Z(G)$ by $Z$. Note that $Z$ is normal by definition, abelian so soluble, and $|G/Z|<|G|$ since $Z$ is not trivial by Theorem \ref{thm:primepowerordergpnontrivZ}. So $G/Z$ is soluble by inductive hypothesis, hence by Lemma \ref{lemma:GsoliffNGNsol} $G$ is soluble.
\end{proof}

\begin{example}[How far can we push this?]
\begin{enumerate}
\item Let $G$ be a group of order $2p^n$. If $p=2$ then $G$ is soluble by theorem above. If $p$ is odd, let $P\in\Syl_p(G)$, then $|G:P|=2$ so $P\unlhd G$. Also, $P$ is soluble by theorem above and $G/P\cong C_2$ is soluble, hence by Lemma \ref{lemma:GsoliffNGNsol} $G$ is soluble.
\begin{flushright}
\textit{Week 10, lecture 3 starts here}
\end{flushright}
\item Say $|G|=4p^n$. Is $G$ soluble? If $p=2$ then again by theorem above $G$ is soluble, so suppose $p$ is odd. Now we know
\begin{itemize}
\item $S_4$ is soluble (see sheet 5 Q3).
\item Let $G$ act on $X=G/P$ where $P\in\Syl_p(G)$ by left multiplication, then $\ker(G,X,\cdot)\leq P$ (see sheet 2 Q9). By Lagrange's, $\ker(G,X,\cdot)$ has $p$-power order, so it's soluble by theorem above.
\item $G/\ker(G,X,\cdot)\cong$ a subgroup of $S_4$ by \ref{prop:GhomSymX}, so it's soluble by \ref{thm:subgpofsolissol}.
\end{itemize}
So by \ref{lemma:GsoliffNGNsol} $G$ is indeed soluble.
\item If $|G|=3p^n$, then $G$ is soluble by the same argument and the fact that $S_3$ is soluble.
\item Challenge: if $|G|=5p^n$, is $G$ soluble? (Yes.) Note $S_5$ is not soluble.
\end{enumerate}
\end{example}

\begin{thm}
Let $G_1,\ldots,G_t$ be finite soluble groups. Then $G=G_1\times\cdots\times G_t$ is soluble.
\end{thm}
\begin{proof}[Proof by induction on t] The base case is a tautology, so suppose $t\geq 2$ and result holds for smaller values of $t$, i.e. $X=G_2\times\cdots G_t$ is soluble and we want to prove $G=G_1\times X$ is as well. Consider the projection homomorphism $\pi:G_1\times X\rightarrow X:(g,x)\mapsto x$. Then $\ker\pi=G_1\times \{1_X\}\cong G_1$ is soluble, and $\im\pi=X$ is also soluble. By the 1st isomorphism theorem, $G/\ker\pi\cong\im\pi$, so by \ref{lemma:GsoliffNGNsol} one has the desired.
\end{proof}

\subsection{Nonexaminable: nilpotent group}
\begin{defn}
For a group $G$, define $\gamma_1(G):=G$ and for $i\geq 2$, $\gamma_i(G):=[\gamma_{i-1}G,G]$. Then $\gamma_1(G)\geq\gamma_2(G)\geq\cdots$ form the \textit{lower central series} of $G$.
\end{defn}
\begin{defn}
A group $G$ is \textit{nilpotent} if $\gamma_n(G)=\{1_G\}$ for some $n\in\N$. The maximal $c\in\N:\gamma_c(G)\neq\{1_G\}$ is called the \textit{nilpotency class} of $G$.
\end{defn}
\begin{example}
\begin{enumerate}
\item All abelian groups are nilpotent of nilpotency class 1.
\item Let $G=D_8=\la \sigma,\tau\ra$. Then $\gamma_2(G)=[G,G]=\la\sigma^2\ra=Z(G)$, and $\gamma_3(G)=[Z(G),G]=\{1_G\}$, so $D_8$ is nilpotent of nilpotency class 2.
\end{enumerate}
\end{example}

\begin{prop}
\begin{enumerate}
\item If $H\leq G$ and $G$ is nilpotent, then $H$ is nilpotent.
\item Let $N\unlhd G$. Then $\gamma_n(G/N)=\gamma_n(G)N/N$. In particular, if $G$ is nilpotent then so is $G/N$.
\end{enumerate}
\end{prop}
\begin{exe}[Before the proof]
Let $A\subseteq G$. Show that $\la A\ra N/N=\la B\ra$ where $B=\{aN:a\in A\}$.
\end{exe}
\begin{proof}
\begin{enumerate}
\item By definition, if $H\leq G$ then $\gamma_n(H)\leq\gamma_n(G) \ \forall n\in N$.
\item We prove by induction on $n$. If $n=1$ then $\gamma_1(G/N)=G/N=GN/N=\gamma_1(G)N/N$, so suppose $n>1$ and result holds for smaller values of $n$. Note $[gN,hN]=[g,h]N \ \forall g,h\in G$. Thus
\[
\begin{aligned}
\gamma_n(G/N)&=\la [xN,yN]:xN\in\gamma_{n-1}(G/N),yN\in G/N\ra \qquad \text{by definition}\\
&=\la [xN,yN]:xN\in\gamma_{n-1}(G)N/N,yN\in G/N \ra \qquad \text{by inductive hypothesis} \\
&=\la[xN,yN]:x\in\gamma_{n-1}(G),y\in G\ra \\
&=\la[x,y]N:x\in\gamma_{n-1}(G),y\in G\ra \\
&=\la[x,y]:x\in\gamma_{n-1}(G),y\in G\ra N/N\qquad\text{by exercise} \\
&=\gamma_n(G)N/N. \qquad\text{by definition again}
\end{aligned}
\]
\end{enumerate}
\end{proof}

\begin{coro}
Every finite nilpotent group is soluble.
\end{coro}
\begin{proof}
First note that $G^{(0)}=G=\gamma_1(G)$ and an easy inductive proof shows more generally
\[
G^{(n-1)}\leq \gamma_n(G) \ \forall n\in \N.
\]
Hence, $G$ nilpotent $\Rightarrow \gamma_n(G)=\{1_G\}$ for some $n\in \N\Rightarrow G^{(n-1)}=\{1_G\}\Rightarrow G$ soluble by Theorem \ref{thm:GsoliffGneq1}.
\end{proof}

\begin{prop}
If $|G|=p^n$ for $p$ prime then $G$ is nilpotent, i.e.
\[
\begin{aligned}
\{\text{finite groups of prime power order}\}&\subseteq\{\text{finite nilpotent groups}\}\\&\subseteq\{\text{finite soluble groups}\}\subseteq\{\text{finite groups}\}.
\end{aligned}
\]
\end{prop}
\end{document}